---
title: "ATRX rescue in an ATRX-KO neuroblastoma line"
subtitle: "Bulk RNA-seq analysis"
author: "Arthur Dondi"
date: "`r format(Sys.Date(), '%d %B %Y')`"
params:
  # Point this at User.output_dir from your config, or pass it at render time.
  outdir: "/nobackup/lab_taschner-mandl/arthurdondi/projects/epicode"
  padj: 0.05
  lfc: 1.0
  # Order groups appear in every figure. Edit if you add conditions.
  condition_levels: !r c("TP53", "E6", "EmptyVector", "ATRX_FL", "ATRX_IFF")
  # The contrast the report leads with.
  headline_contrast: "ATRX_FL_vs_ATRX_IFF"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    self_contained: true
    theme: flatly
    fig_width: 9
    fig_height: 5
    df_print: kable
  pdf_document:
    toc: true
    number_sections: true
    fig_width: 8
    fig_height: 4.5
---

<!--
HOW TO RENDER
  Rscript -e 'rmarkdown::render("report/analysis_report.Rmd",
              params = list(outdir = "/path/to/output_dir"),
              output_file = "analysis_report.html")'

  Swap output_format = "pdf_document" for PDF (needs LaTeX).

HOW THIS WORKS
  Every figure is rebuilt here from the CSV files the pipeline already writes,
  rather than embedding the PDFs. That keeps the styling consistent, makes the
  HTML self-contained, and means you can edit any plot in this file.

  Every section degrades gracefully: if an output is missing the section says so
  and the report still renders. So you can run it before the pipeline finishes.

  Text in > blockquotes is placeholder interpretation - rewrite it.
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      dpi = 150, fig.align = "center")

suppressPackageStartupMessages({
  library(ggplot2); library(dplyr); library(tidyr); library(knitr)
})

OUT <- params$outdir
LEV <- params$condition_levels

# Condition palette: the ggplot hue values, so figures match the pipeline PDFs.
PAL <- setNames(
  c("#E76BF3", "#00BF7D", "#00B0F6", "#F8766D", "#A3A500")[seq_along(LEV)], LEV)

theme_set(theme_bw(base_size = 11))
theme_update(panel.grid.minor = element_blank(),
             plot.title = element_text(face = "bold", size = 12),
             plot.subtitle = element_text(colour = "grey30", size = 9))

# ── Helpers ──────────────────────────────────────────────────────────────────
# Read a pipeline output, or return NULL and leave a visible note in the report.
grab <- function(...) {
  f <- file.path(OUT, ...)
  if (!file.exists(f)) return(NULL)
  read.csv(f, stringsAsFactors = FALSE, check.names = FALSE)
}
missing_note <- function(...) {
  cat("\n> **Not available.** `", file.path(...), "` was not found. ",
      "Run that part of the pipeline and re-render.\n\n", sep = "")
}
cond_factor <- function(x) factor(x, levels = LEV[LEV %in% unique(x)])

# Long-format helper for the wide per-gene tables (gene_id + one column per sample).
gene_long <- function(df, meta) {
  if (is.null(df) || nrow(df) == 0) return(NULL)
  df %>%
    pivot_longer(-gene_id, names_to = "sample", values_to = "value") %>%
    left_join(meta, by = "sample") %>%
    mutate(condition = cond_factor(condition),
           sample = factor(sample, levels = meta$sample))
}

# Sample sheet, used to colour everything by condition.
qc <- grab("design_qc", "design_qc_summary.csv")
meta <- if (!is.null(qc)) {
  qc %>% select(sample, condition) %>%
    arrange(match(condition, LEV)) %>%
    mutate(condition = cond_factor(condition))
} else NULL

volcano <- function(res, title, num, den) {
  res <- res %>% filter(!is.na(padj)) %>%
    mutate(sig = padj < params$padj & abs(log2FoldChange) >= params$lfc)
  ggplot(res, aes(log2FoldChange, -log10(padj), colour = sig)) +
    geom_point(size = 0.8, alpha = 0.5) +
    geom_vline(xintercept = c(-params$lfc, params$lfc), linetype = "dashed",
               linewidth = 0.3) +
    geom_hline(yintercept = -log10(params$padj), linetype = "dashed",
               linewidth = 0.3) +
    scale_colour_manual(values = c("FALSE" = "grey70", "TRUE" = "#B2182B"),
                        guide = "none") +
    labs(title = title,
         subtitle = sprintf("%s significant | right = up in %s, left = up in %s",
                            sum(res$sig), num, den),
         x = sprintf("log2FC (%s / %s)", num, den),
         y = expression(-log[10] ~ "adjusted p"))
}

# Ranking by padj alone floats high-count genes with tiny fold-changes to the
# top. This ranks significant genes by -log10(padj) * |log2FC| so both matter,
# and reports the two directions separately so one cannot crowd out the other.
top_genes <- function(res, n = 10) {
  res %>% filter(!is.na(padj), padj < params$padj,
                 abs(log2FoldChange) >= params$lfc) %>%
    mutate(direction = ifelse(log2FoldChange > 0, "up", "down"),
           score = -log10(padj) * abs(log2FoldChange)) %>%
    group_by(direction) %>% slice_max(score, n = n) %>% ungroup() %>%
    arrange(direction, desc(score)) %>%
    transmute(gene_id, direction, baseMean = round(baseMean),
              log2FC = round(log2FoldChange, 2), padj = signif(padj, 3),
              score = round(score, 1))
}

de_counts <- function(res) {
  r <- res %>% filter(!is.na(padj), padj < params$padj,
                      abs(log2FoldChange) >= params$lfc)
  c(up = sum(r$log2FoldChange > 0), down = sum(r$log2FoldChange < 0))
}
```

# Experimental design

```{r design}
data.frame(
  Group = c("TP53", "E6", "EmptyVector", "ATRX_FL", "ATRX_IFF"),
  Description = c("TP53 KO", "TP53 KO + ATRX KO",
                  "E6 + empty EGFP/mCherry vector",
                  "E6 + ATRX full length (rescue)",
                  "E6 + ATRX in-frame fusion (semi-rescue)"),
  `Parental clone` = c("TP53", "E6", "E6", "E6", "E6"),
  Vector = c("none", "none", "empty", "ATRX-FL", "ATRX-IFF"),
  n = c(3, 3, 3, 3, 3), check.names = FALSE
) %>% kable()
```

> **Read this before the figures.** Every group is a single clone, and each
> construct is a single transduced pool. The TP53 line and the E6 lineage differ
> by clonal background as well as by ATRX genotype, and those two cannot be
> separated in this design. Comparisons **within** the E6 lineage
> (`ATRX_FL` / `ATRX_IFF` / `EmptyVector`) are internally controlled and are the
> ones to trust; comparisons **against TP53** carry the clonal difference and are
> shown for completeness only.

---

# Quality control

## Library-level metrics

```{r qc-lib, fig.height = 4}
if (is.null(qc)) missing_note("design_qc", "design_qc_summary.csv") else {
  d <- qc %>% mutate(condition = cond_factor(condition),
                     sample = factor(sample, levels = meta$sample))
  print(
    d %>% select(sample, condition, assigned_reads, genes_detected, pct_assigned) %>%
      pivot_longer(c(assigned_reads, genes_detected, pct_assigned)) %>%
      mutate(name = recode(name, assigned_reads = "Reads assigned to genes",
                           genes_detected = "Genes detected",
                           pct_assigned = "% reads assigned")) %>%
      ggplot(aes(sample, value, fill = condition)) +
      geom_col() +
      facet_wrap(~ name, scales = "free_y", ncol = 1) +
      scale_fill_manual(values = PAL) +
      labs(x = NULL, y = NULL, fill = "Condition") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

> No group-linked shift in depth, complexity or assignment rate — the libraries
> are comparable. *(Check and rewrite.)*

## Sample–sample correlation

```{r qc-cor, fig.height = 6}
cm <- grab("design_qc", "sample_correlation.csv")
if (is.null(cm)) missing_note("design_qc", "sample_correlation.csv") else {
  m <- as.matrix(cm[, -1]); rownames(m) <- cm$sample
  ord <- meta$sample[meta$sample %in% rownames(m)]
  m <- m[ord, ord]
  print(
    as.data.frame(as.table(m)) %>%
      setNames(c("x", "y", "rho")) %>%
      mutate(x = factor(x, ord), y = factor(y, rev(ord))) %>%
      ggplot(aes(x, y, fill = rho)) +
      geom_tile(colour = "white", linewidth = 0.3) +
      scale_fill_gradient(low = "white", high = "#2166AC") +
      coord_fixed() +
      labs(title = "Spearman correlation (VST)", x = NULL, y = NULL) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            panel.grid = element_blank())
  )
}
```

> Replicates form tight blocks on the diagonal. *(Check and rewrite.)*

## ATRX and vector markers

Reporter and transgene features are rows in the count matrix because the
reference includes the vector contigs. They confirm the constructs are where
they should be, and nowhere else.

```{r qc-markers, fig.height = 8}
mk <- gene_long(grab("design_qc", "marker_expression.csv"), meta)
if (is.null(mk)) missing_note("design_qc", "marker_expression.csv") else {
  print(
    ggplot(mk, aes(sample, value + 1, fill = condition)) +
      geom_col() +
      facet_wrap(~ gene_id, scales = "free_y", ncol = 1) +
      scale_y_log10() + scale_fill_manual(values = PAL) +
      labs(title = "Reporter and transgene expression",
           subtitle = "normalised counts + 1, log10; these features are excluded from the size factors",
           x = NULL, y = "normalised count + 1", fill = "Condition") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

```{r qc-atrx-table}
mk2 <- grab("design_qc", "marker_expression.csv")
if (!is.null(mk2) && "ATRX" %in% mk2$gene_id && !is.null(meta)) {
  gene_long(mk2, meta) %>% filter(gene_id == "ATRX") %>%
    group_by(condition) %>%
    summarise(mean = round(mean(value)), min = round(min(value)),
              max = round(max(value)), .groups = "drop") %>%
    mutate(`% of TP53` = round(100 * mean / mean[condition == "TP53"]),
           `fold spread` = sprintf("%.1fx", max / pmax(min, 1))) %>%
    kable(caption = "Endogenous ATRX locus, normalised counts")
}
```

> EGFP/mCherry are present only in the transduced groups; each ATRX construct is
> detected only in its own group. *(Add your read of the ATRX levels: is the
> rescue at a physiological dose? Is one construct variable across replicates?)*

---

# Global structure (PCA)

```{r pca, fig.height = 5, results = "asis"}
pca_views <- list(
  c("all", "no_vector_genes", "All samples, reporter features removed"),
  c("E6_derived", "no_vector_no_transgene",
    "E6-derived only, reporter + ATRX removed")
)
for (v in pca_views) {
  co <- grab("pca", v[1], paste0(v[2], "_coords.csv"))
  ve <- grab("pca", v[1], paste0(v[2], "_variance_explained.csv"))
  cat("\n\n##", v[3], "\n\n")
  if (is.null(co)) { missing_note("pca", v[1], paste0(v[2], "_coords.csv")); next }
  pv <- if (!is.null(ve)) setNames(ve$percent_variance, ve$PC) else c(PC1 = NA, PC2 = NA)
  p <- co %>% mutate(condition = cond_factor(condition)) %>%
    ggplot(aes(PC1, PC2, colour = condition, label = sample)) +
    geom_point(size = 3) +
    ggrepel::geom_text_repel(size = 2.6, show.legend = FALSE, max.overlaps = 20) +
    scale_colour_manual(values = PAL) +
    labs(title = v[3],
         x = sprintf("PC1: %.1f%% variance", pv[["PC1"]]),
         y = sprintf("PC2: %.1f%% variance", pv[["PC2"]]),
         colour = "Condition")
  print(p)
}
```

> PC1 of the full PCA separates TP53 from everything E6-derived: that is the
> clonal background, confounded with the ATRX genotype and not correctable.
> Restricting to the E6 lineage lets the construct effects use PC1/PC2.
> *(Rewrite with what you see.)*

---

# How TP53 differs from the E6 lineage

Before narrowing to the constructs, this is the size of the difference we are
setting aside. `TP53` is a different clone from everything E6-derived, and that
difference is confounded with the ATRX genotype: no comparison can separate them.

```{r tp53-vs-e6, fig.height = 5}
r_tp53 <- grab("deseq2", "TP53_vs_E6", "results.csv")
if (is.null(r_tp53)) missing_note("deseq2", "TP53_vs_E6", "results.csv") else {
  print(volcano(r_tp53, "TP53_vs_E6", num = "E6", den = "TP53"))
}
```

```{r tp53-scale}
if (!is.null(r_tp53)) {
  n <- de_counts(r_tp53)
  cat(sprintf(paste("\n> **%s genes** change between TP53 and E6 at padj < %s and",
                    "|log2FC| >= %s (%s up in E6, %s down) - roughly %.0f%% of the",
                    "genes tested. That is the clone plus the ATRX knockout",
                    "together, and it dwarfs everything that follows.\n\n"),
              format(sum(n), big.mark = ","), params$padj, params$lfc,
              format(n["up"], big.mark = ","), format(n["down"], big.mark = ","),
              100 * sum(n) / sum(!is.na(r_tp53$padj))))
}
```

```{r tp53-top}
if (!is.null(r_tp53)) kable(top_genes(r_tp53), caption = "TP53_vs_E6: top genes")
```

> *(Say what the TP53-vs-E6 difference looks like biologically, and state
> explicitly that from here on TP53 is set aside because the clonal background
> cannot be separated from the ATRX genotype.)*

---

# Differential expression

## `r params$headline_contrast` — the primary comparison

Both groups share the clone and carry a vector, so the clonal and transduction
effects cancel. This is the cleanest contrast in the dataset.

```{r de-headline, fig.height = 5}
hc  <- params$headline_contrast
res <- grab("deseq2", hc, "results.csv")
if (is.null(res)) missing_note("deseq2", hc, "results.csv") else {
  # EDIT these two if you change the headline contrast.
  print(volcano(res, hc, num = "ATRX_IFF", den = "ATRX_FL"))
}
```

```{r de-headline-top}
if (!is.null(res)) {
  kable(top_genes(res),
        caption = paste("Top 10 up and 10 down, ranked by -log10(padj) x |log2FC|",
                        "among genes passing both thresholds"))
}
```

> *(Your interpretation of the FL vs IFF difference.)*

## Genes of interest

The point of this panel is the **pattern across groups**, not any single gene.

```{r goi, fig.height = 8}
goi <- gene_long(grab("design_qc", "goi_expression.csv"), meta)
if (is.null(goi)) missing_note("design_qc", "goi_expression.csv") else {
  print(
    ggplot(goi, aes(sample, value + 1, fill = condition)) +
      geom_col() +
      facet_wrap(~ gene_id, scales = "free_y", ncol = 1) +
      scale_y_log10() + scale_fill_manual(values = PAL) +
      labs(title = "Genes of interest", subtitle = "normalised counts + 1, log10",
           x = NULL, y = "normalised count + 1", fill = "Condition") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

```{r goi-means, fig.height = 4}
if (!is.null(goi)) {
  print(
    goi %>% group_by(gene_id, condition) %>%
      summarise(mean = mean(value), sd = sd(value), .groups = "drop") %>%
      ggplot(aes(condition, mean, fill = condition)) +
      geom_col() +
      geom_errorbar(aes(ymin = pmax(mean - sd, 0), ymax = mean + sd), width = 0.25,
                    linewidth = 0.3) +
      facet_wrap(~ gene_id, scales = "free_y", nrow = 1) +
      scale_fill_manual(values = PAL, guide = "none") +
      labs(title = "Group means (± SD)", x = NULL, y = "normalised count") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

> *(Describe the pattern: which groups move together, and where TP53 sits
> relative to them. Note that TP53 is a different clone, so a difference there is
> not by itself evidence about ATRX.)*

## Remaining contrasts

```{r de-others, fig.height = 4.5, results = "asis"}
# EDIT: add or reorder. Each entry is c(contrast_dir, numerator, denominator).
contrasts <- list(
  c("EmptyVector_vs_ATRX_FL",  "ATRX_FL",     "EmptyVector"),
  c("EmptyVector_vs_ATRX_IFF", "ATRX_IFF",    "EmptyVector"),
  c("E6_vs_EmptyVector",       "EmptyVector", "E6"),
  # TP53_vs_E6 has its own section above.
  c("TP53_vs_ATRX_FL",         "ATRX_FL",     "TP53"),
  c("TP53_vs_ATRX_IFF",        "ATRX_IFF",    "TP53")
)
tally <- list()
for (c3 in contrasts) {
  r <- grab("deseq2", c3[1], "results.csv")
  cat("\n\n###", c3[1], "\n\n")
  if (is.null(r)) { missing_note("deseq2", c3[1], "results.csv"); next }
  tally[[c3[1]]] <- de_counts(r)
  print(volcano(r, c3[1], num = c3[2], den = c3[3]))
}
```

```{r de-tally}
if (length(tally)) {
  do.call(rbind, tally) %>% as.data.frame() %>%
    tibble::rownames_to_column("contrast") %>%
    mutate(total = up + down) %>%
    kable(caption = sprintf("Genes with padj < %s and |log2FC| >= %s",
                            params$padj, params$lfc))
}
```

---

# Pathway enrichment (GSEA)

```{r gsea, fig.height = 6, results = "asis"}
gsea_dot <- function(df, title, top = 12) {
  df <- df %>% filter(!is.na(padj)) %>% arrange(desc(abs(NES)))
  df <- bind_rows(head(filter(df, NES > 0), top), head(filter(df, NES < 0), top))
  if (!nrow(df)) return(invisible(NULL))
  df <- df %>% mutate(
    pathway = gsub("^HALLMARK_", "", pathway),
    pathway = factor(pathway, levels = pathway[order(df$NES)]),
    side = ifelse(NES > 0, paste("Enriched in", df$numerator[1]),
                  paste("Enriched in", df$denominator[1])))
  print(
    ggplot(df, aes(NES, pathway, colour = padj, size = size)) +
      geom_point() + geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3) +
      scale_colour_gradient(low = "#B2182B", high = "grey75", limits = c(0, 0.25)) +
      labs(title = title, y = NULL, colour = "padj", size = "Set size")
  )
}
# EDIT: which contrasts to show enrichment for.
for (cn in c("ATRX_FL_vs_ATRX_IFF", "EmptyVector_vs_ATRX_FL",
             "EmptyVector_vs_ATRX_IFF")) {
  g <- grab("gsea", cn, "H_results.csv")
  cat("\n\n##", cn, "\n\n")
  if (is.null(g)) { missing_note("gsea", cn, "H_results.csv"); next }
  gsea_dot(g, paste("Hallmark:", cn))
}
```

> Hallmark sets overlap heavily, and the proliferation sets (E2F, G2M, MYC) and
> the inflammatory/interferon sets are two ends of one axis that dominates most
> cell-line RNA-seq. Read them as one signal, not as many independent findings.
> *(Rewrite with your conclusion.)*

---

# Rescue assessment

Whether the rescue worked is **not** answerable by asking if `ATRX_FL` clusters
with `TP53` — different clone, different transduction status. Instead: does the
intervention push the knockout-responsive genes back the other way? Both
contrasts used here sit inside a single genetic background, so the clonal
difference never enters.

## Signature reversal

```{r rescue-stats}
srs <- list.files(file.path(OUT, "signature_reversal"), pattern = "reversal_stats.csv",
                  recursive = TRUE, full.names = TRUE)
if (!length(srs)) missing_note("signature_reversal", "*", "reversal_stats.csv") else {
  # bind_rows, not rbind: stats files written by different versions of the script
  # can have different column sets, and rbind would abort the whole render.
  tab <- bind_rows(lapply(srs, read.csv)) %>%
    select(any_of(c("comparison", "n_signature_genes", "pct_reversed",
                    "null_pct_reversed", "excess_over_null", "perm_p",
                    "spearman_rho", "interpretable")))
  if ("excess_over_null" %in% names(tab)) tab <- arrange(tab, desc(excess_over_null))
  stale <- setdiff(c("null_pct_reversed", "excess_over_null", "perm_p"), names(tab))
  if (length(stale)) {
    cat("\n> **Note:** some `reversal_stats.csv` files predate the permutation",
        "null and lack", paste0("`", stale, "`", collapse = ", "),
        "- re-run those comparisons.\n\n")
  }
  kable(tab, caption = paste("Reversal of the knockout signature. Read the excess",
                             "over the permutation null, not the raw percentage."))
}
```

```{r rescue-scatter, fig.height = 6, results = "asis"}
# EDIT: which comparison to draw in full.
for (nm in c("ATRX_FL_reverses_KO_signature_lfc1",
             "ATRX_IFF_reverses_KO_signature_lfc1")) {
  g <- grab("signature_reversal", nm, "reversal_genes.csv")
  cat("\n\n###", nm, "\n\n")
  if (is.null(g)) { missing_note("signature_reversal", nm, "reversal_genes.csv"); next }
  lim <- max(abs(c(g$lfc_sig, g$lfc_resp)), na.rm = TRUE)
  st  <- grab("signature_reversal", nm, "reversal_stats.csv")
  # Stats go inside the panel: a long subtitle is silently clipped once the
  # figure is embedded or scaled, and these numbers are the point of the plot.
  box <- if (!is.null(st)) paste(
    sprintf("slope     %+.2f", st$tls_slope[1]),
    sprintf("R\u00b2        %.3f", if ("r_squared" %in% names(st))
              st$r_squared[1] else st$pearson_r[1]^2),
    sprintf("rho       %+.2f", st$spearman_rho[1]),
    sprintf("reversed  %.0f%% (null %.0f%%)", st$pct_reversed[1],
            st$null_pct_reversed[1]),
    sprintf("perm p    %s", format(signif(st$perm_p[1], 2))), sep = "\n") else NULL
  p <- ggplot(g, aes(lfc_sig, lfc_resp)) +
    annotate("rect", xmin = 0, xmax = lim, ymin = -lim, ymax = 0,
             fill = "#4DAF4A", alpha = .07) +
    annotate("rect", xmin = -lim, xmax = 0, ymin = 0, ymax = lim,
             fill = "#4DAF4A", alpha = .07) +
    geom_hline(yintercept = 0, colour = "grey60", linewidth = .3) +
    geom_vline(xintercept = 0, colour = "grey60", linewidth = .3) +
    geom_abline(slope = -1, linetype = "dashed", colour = "grey45") +
    geom_point(size = 1, alpha = .5, colour = "#2166AC") +
    coord_fixed(xlim = c(-lim, lim), ylim = c(-lim, lim)) +
    labs(title = nm,
         subtitle = "shaded = reversal | dashed line = complete reversal (slope -1)",
         x = "log2FC knockout (the damage)",
         y = "log2FC rescue (the repair)")
  if (!is.null(box)) {
    p <- p + annotate("label", x = lim, y = lim, hjust = 1, vjust = 1, label = box,
                      size = 2.9, family = "mono", lineheight = 1.15,
                      fill = grDevices::adjustcolor("white", alpha.f = .88),
                      label.size = .25, colour = "grey15")
  }
  print(p)
}
```

> A gene the knockout pushed up should be pushed back down: points in the shaded
> quadrants, along the dashed line. Slope −1 is a complete rescue, 0 is no
> response. *(Rewrite with the numbers from the table above.)*

## Interaction contrasts

`(rescue / EmptyVector) − (TP53 / E6)`: how far the rescue moved each gene minus
how far it needed to move. Zero is a complete rescue; negative is a shortfall.

```{r rescue-interaction, fig.height = 4.5, results = "asis"}
for (nm in c("ATRX_FL_rescue_shortfall", "ATRX_IFF_rescue_shortfall")) {
  r <- grab("deseq2_interaction", nm, "results.csv")
  cat("\n\n###", nm, "\n\n")
  if (is.null(r)) { missing_note("deseq2_interaction", nm, "results.csv"); next }
  print(volcano(r, nm, num = "overshoot", den = "shortfall"))
}
```

---

# The E6 lineage alone

With `TP53` excluded, every remaining group shares one clone and one
transduction step. Differences here are attributable to the construct — the two
effects we could not separate above are now held constant by design rather than
corrected for.

## PCA, E6-derived samples only

```{r e6-pca, fig.height = 5}
co <- grab("pca", "E6_derived", "no_vector_no_transgene_coords.csv")
ve <- grab("pca", "E6_derived", "no_vector_no_transgene_variance_explained.csv")
if (is.null(co)) {
  missing_note("pca", "E6_derived", "no_vector_no_transgene_coords.csv")
} else {
  pv <- if (!is.null(ve)) setNames(ve$percent_variance, ve$PC) else c(PC1=NA, PC2=NA)
  print(
    co %>% mutate(condition = cond_factor(condition)) %>%
      ggplot(aes(PC1, PC2, colour = condition, label = sample)) +
      geom_point(size = 3) +
      ggrepel::geom_text_repel(size = 2.6, show.legend = FALSE, max.overlaps = 20) +
      scale_colour_manual(values = PAL) +
      labs(title = "E6-derived samples, reporter and ATRX features removed",
           subtitle = "clone and transduction held constant; the construct is the only intended difference",
           x = sprintf("PC1: %.1f%% variance", pv[["PC1"]]),
           y = sprintf("PC2: %.1f%% variance", pv[["PC2"]]),
           colour = "Condition")
  )
}
```

## Sample–sample correlation, E6-derived only

Pairwise correlations are unchanged by dropping TP53 — but the colour scale
rescales to the range that remains, so differences compressed by the clonal
outlier become visible.

```{r e6-cor, fig.height = 5.5}
if (is.null(cm)) missing_note("design_qc", "sample_correlation.csv") else {
  m <- as.matrix(cm[, -1]); rownames(m) <- cm$sample
  keep <- meta$sample[meta$condition != "TP53" & meta$sample %in% rownames(m)]
  if (length(keep) < 3) {
    cat("\n> Fewer than 3 non-TP53 samples — nothing to plot.\n\n")
  } else {
    m <- m[keep, keep]
    print(
      as.data.frame(as.table(m)) %>%
        setNames(c("x", "y", "rho")) %>%
        mutate(x = factor(x, keep), y = factor(y, rev(keep))) %>%
        ggplot(aes(x, y, fill = rho)) +
        geom_tile(colour = "white", linewidth = 0.3) +
        geom_text(aes(label = sprintf("%.3f", rho)), size = 2.1, colour = "grey20") +
        scale_fill_gradient(low = "white", high = "#2166AC") +
        coord_fixed() +
        labs(title = "Spearman correlation (VST), TP53 excluded",
             subtitle = "scale rescaled to the E6 lineage",
             x = NULL, y = NULL) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
              panel.grid = element_blank())
    )
  }
}
```

> *(Do the replicates of each construct still form the tightest blocks once the
> clonal outlier is removed? If a construct's replicates correlate no better with
> each other than with another construct's, treat its contrasts cautiously.)*

## The three construct comparisons

```{r e6-three, fig.height = 4.5, results = "asis"}
trio <- list(
  c("EmptyVector_vs_ATRX_FL",  "ATRX_FL",  "EmptyVector",
    "Full-length rescue against its own transduced control"),
  c("EmptyVector_vs_ATRX_IFF", "ATRX_IFF", "EmptyVector",
    "In-frame fusion against the same control"),
  c("ATRX_FL_vs_ATRX_IFF",     "ATRX_IFF", "ATRX_FL",
    "The two constructs against each other")
)
trio_tally <- list()
for (t4 in trio) {
  r <- grab("deseq2", t4[1], "results.csv")
  cat("\n\n###", t4[4], "\n\n")
  if (is.null(r)) { missing_note("deseq2", t4[1], "results.csv"); next }
  trio_tally[[t4[1]]] <- de_counts(r)
  print(volcano(r, t4[1], num = t4[2], den = t4[3]))
  cat("\n\n")
  print(kable(top_genes(r, n = 8)))
  cat("\n\n")
}
```

```{r e6-trio-tally}
if (length(trio_tally)) {
  do.call(rbind, trio_tally) %>% as.data.frame() %>%
    tibble::rownames_to_column("contrast") %>% mutate(total = up + down) %>%
    kable(caption = "Differentially expressed genes within the E6 lineage")
}
```

## Genes of interest, E6 lineage only

```{r e6-goi, fig.height = 4}
if (!is.null(goi)) {
  print(
    goi %>% filter(condition != "TP53") %>%
      group_by(gene_id, condition) %>%
      summarise(mean = mean(value), sd = sd(value), .groups = "drop") %>%
      ggplot(aes(condition, mean, fill = condition)) +
      geom_col() +
      geom_errorbar(aes(ymin = pmax(mean - sd, 0), ymax = mean + sd),
                    width = .25, linewidth = .3) +
      facet_wrap(~ gene_id, scales = "free_y", nrow = 1) +
      scale_fill_manual(values = PAL, guide = "none") +
      labs(title = "Genes of interest, TP53 excluded", x = NULL,
           y = "normalised count") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

> *(This is the section your conclusion should rest on. Which construct moves
> the genes of interest, in which direction, and by how much — with everything
> that could confound it held constant rather than corrected.)*

---

# Summary

> *(Three or four sentences. What the rescue did, what it did not, and the one
> caveat a reader must keep: one clone and one transduced pool per condition, so
> construct effects and pool effects cannot be fully separated.)*

```{r session, results = "asis"}
cat("\n<details><summary>Session info</summary>\n\n```\n")
cat(capture.output(sessionInfo()), sep = "\n")
cat("\n```\n</details>\n")
```
