---
title: "Epicode: ATRX rescue in an ATRX-KO neuroblastoma line"
subtitle: "Bulk RNA-seq analysis"
author: "Arthur Dondi"
date: "`r format(Sys.Date(), '%d %B %Y')`"
params:
  # Point this at User.output_dir from your config, or pass it at render time.
  outdir: "/nobackup/lab_taschner-mandl/arthurdondi/projects/epicode"
  padj: 0.05
  lfc: 1.0
  # Order groups appear in every figure. Edit if you add conditions.
  condition_levels: !r c("TP53", "E6", "EmptyVector", "ATRX_FL", "ATRX_IFF")
  # The contrast the report leads with.
  headline_contrast: "ATRX_FL_vs_ATRX_IFF"
  # Vector and transgene features. They are rows in the count matrix because the
  # reference carries the vector contigs, but they are constructs rather than
  # biology, so they are dropped from every differential-expression figure and
  # table below. Removed here in the report only - the pipeline's own outputs
  # still contain them.
  exclude_genes: !r c("ATRX", "ATRX_FL", "ATRX_IFF", "EGFP", "mCherry")
  # Labelled on the headline volcano.
  highlight_genes: !r c("HOXB2", "HOXB4", "HOXB6", "HOXB7", "IGF2BP1")
  # Every volcano additionally labels its N most significant genes (by
  # -log10(padj) x |log2FC|, the same ranking as the top-genes tables), the way
  # the pipeline's own volcano.pdf labels every significant gene. The report
  # caps it at a fixed N instead, since some contrasts have thousands of
  # significant genes and labelling all of them is unreadable at report scale.
  volcano_label_top: 12
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    self_contained: true
    theme: flatly
    fig_width: 9
    fig_height: 5
    df_print: kable
    # Every chunk's code is included but folded behind a "Show" toggle, so the
    # report reads as figures and tables by default while staying fully
    # reproducible one click away. Requires echo = TRUE in the setup chunk below
    # - code_folding only folds code that would otherwise be printed.
    code_folding: hide
  pdf_document:
    toc: true
    number_sections: true
    fig_width: 8
    fig_height: 4.5
---

<!--
HOW TO RENDER
  Rscript -e 'rmarkdown::render("report/analysis_report.Rmd",
              params = list(outdir = "/path/to/output_dir"),
              output_file = "analysis_report.html")'

  Swap output_format = "pdf_document" for PDF (needs LaTeX).

HOW THIS WORKS
  Every figure is rebuilt here from the CSV files the pipeline already writes,
  rather than embedding the PDFs. That keeps the styling consistent, makes the
  HTML self-contained, and means you can edit any plot in this file.

  Every section degrades gracefully: if an output is missing the section says so
  and the report still renders. So you can run it before the pipeline finishes.

  Text in > blockquotes is placeholder interpretation - rewrite it.
-->

```{r setup, include = FALSE}
# echo = TRUE + code_folding: hide (in the YAML header) is the "Show" toggle
# per chunk - code is present in the HTML but folded away by default.
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      dpi = 150, fig.align = "center")

suppressPackageStartupMessages({
  library(ggplot2); library(dplyr); library(tidyr); library(knitr)
})

OUT <- params$outdir
LEV <- params$condition_levels

# Condition palette: the ggplot hue values, so figures match the pipeline PDFs.
PAL <- setNames(
  c("#E76BF3", "#00BF7D", "#00B0F6", "#F8766D", "#A3A500")[seq_along(LEV)], LEV)

theme_set(theme_bw(base_size = 11))
theme_update(panel.grid.minor = element_blank(),
             plot.title = element_text(face = "bold", size = 12),
             plot.subtitle = element_text(colour = "grey30", size = 9))

# ── Helpers ──────────────────────────────────────────────────────────────────
# Read a pipeline output, or return NULL and leave a visible note in the report.
grab <- function(...) {
  f <- file.path(OUT, ...)
  if (!file.exists(f)) return(NULL)
  read.csv(f, stringsAsFactors = FALSE, check.names = FALSE)
}
missing_note <- function(...) {
  cat("\n> **Not available.** `", file.path(...), "` was not found. ",
      "Run that part of the pipeline and re-render.\n\n", sep = "")
}
cond_factor <- function(x) factor(x, levels = LEV[LEV %in% unique(x)])

# Long-format helper for the wide per-gene tables (gene_id + one column per sample).
gene_long <- function(df, meta) {
  if (is.null(df) || nrow(df) == 0) return(NULL)
  df %>%
    pivot_longer(-gene_id, names_to = "sample", values_to = "value") %>%
    left_join(meta, by = "sample") %>%
    mutate(condition = cond_factor(condition),
           sample = factor(sample, levels = meta$sample))
}

# Sample sheet, used to colour everything by condition.
qc <- grab("design_qc", "design_qc_summary.csv")
meta <- if (!is.null(qc)) {
  qc %>% select(sample, condition) %>%
    arrange(match(condition, LEV)) %>%
    mutate(condition = cond_factor(condition))
} else NULL

# DESeq results with the vector/transgene features dropped. They are guaranteed
# to be the most significant genes in any contrast that crosses a construct
# boundary, and they say nothing about the cells' biology.
de_results <- function(...) {
  r <- grab(...)
  if (is.null(r)) return(NULL)
  n0 <- nrow(r)
  r  <- r[!tolower(r$gene_id) %in% tolower(params$exclude_genes), , drop = FALSE]
  attr(r, "n_dropped") <- n0 - nrow(r)
  r
}

volcano <- function(res, title, num, den, highlight = NULL,
                    label_top = params$volcano_label_top) {
  res <- res %>% filter(!is.na(padj)) %>%
    mutate(sig = padj < params$padj & abs(log2FoldChange) >= params$lfc)
  n <- c(up = sum(res$sig & res$log2FoldChange > 0),
         down = sum(res$sig & res$log2FoldChange < 0))
  p <- ggplot(res, aes(log2FoldChange, -log10(padj), colour = sig)) +
    geom_point(size = 0.8, alpha = 0.5) +
    geom_vline(xintercept = c(-params$lfc, params$lfc), linetype = "dashed",
               linewidth = 0.3) +
    geom_hline(yintercept = -log10(params$padj), linetype = "dashed",
               linewidth = 0.3) +
    scale_colour_manual(values = c("FALSE" = "grey70", "TRUE" = "#B2182B"),
                        guide = "none") +
    labs(
      title = title,
      # Spell the direction out on both axes: "up in X" is ambiguous unless the
      # reader already knows which group is the numerator.
      subtitle = sprintf(
        paste0("%s significant (padj < %s, |log2FC| >= %s)\n",
               "RIGHT, log2FC > 0: %s genes HIGHER in %s   |   ",
               "LEFT, log2FC < 0: %s genes HIGHER in %s"),
        format(sum(res$sig), big.mark = ","), params$padj, params$lfc,
        format(n["up"], big.mark = ","), num,
        format(n["down"], big.mark = ","), den),
      x = sprintf("log2FC  (%s / %s)   \u2014   positive = higher in %s", num, den, num),
      y = expression(-log[10] ~ "adjusted p")) +
    theme(plot.subtitle = element_text(lineheight = 1.2))
  # Label the most significant genes, ranked the same way as top_genes():
  # -log10(padj) x |log2FC|, so the labelled genes match the accompanying table.
  # The pipeline's own volcano.pdf labels every significant gene; capped here
  # because report-scale figures become unreadable once a contrast has more
  # than a few dozen significant genes (TP53_vs_E6 has thousands).
  hl_lower <- if (is.null(highlight)) character(0) else tolower(highlight)
  if (label_top > 0) {
    lbl <- res %>% filter(sig, !tolower(gene_id) %in% hl_lower) %>%
      mutate(score = -log10(padj) * abs(log2FoldChange)) %>%
      slice_max(score, n = label_top, with_ties = FALSE)
    if (nrow(lbl)) {
      p <- p + ggrepel::geom_text_repel(
        data = lbl, aes(label = gene_id), colour = "grey20", size = 2.4,
        max.overlaps = 30, segment.size = 0.2, segment.colour = "grey50",
        show.legend = FALSE)
    }
  }
  if (!is.null(highlight)) {
    h <- res %>% filter(tolower(gene_id) %in% hl_lower)
    if (nrow(h)) {
      # Drawn after the general labels so the highlighted genes and their
      # labels sit on top rather than being obscured by them.
      p <- p +
        geom_point(data = h, colour = "#1B7837", size = 2.4) +
        ggrepel::geom_text_repel(
          data = h, aes(label = gene_id), colour = "#1B7837", fontface = "bold",
          size = 3.1, min.segment.length = 0, box.padding = 0.6,
          segment.colour = "#1B7837", max.overlaps = Inf, show.legend = FALSE)
    }
    miss <- setdiff(hl_lower, tolower(res$gene_id))
    if (length(miss)) message("highlight not found: ", paste(miss, collapse = ", "))
  }
  p
}

# Ranking by padj alone floats high-count genes with tiny fold-changes to the
# top. This ranks significant genes by -log10(padj) * |log2FC| so both matter,
# and reports the two directions separately so one cannot crowd out the other.
top_genes <- function(res, n = 10) {
  res %>% filter(!is.na(padj), padj < params$padj,
                 abs(log2FoldChange) >= params$lfc) %>%
    mutate(direction = ifelse(log2FoldChange > 0, "up", "down"),
           score = -log10(padj) * abs(log2FoldChange)) %>%
    group_by(direction) %>% slice_max(score, n = n) %>% ungroup() %>%
    arrange(direction, desc(score)) %>%
    transmute(gene_id, direction, baseMean = round(baseMean),
              log2FC = round(log2FoldChange, 2), padj = signif(padj, 3),
              score = round(score, 1))
}

de_counts <- function(res) {
  r <- res %>% filter(!is.na(padj), padj < params$padj,
                      abs(log2FoldChange) >= params$lfc)
  c(up = sum(r$log2FoldChange > 0), down = sum(r$log2FoldChange < 0))
}
```

# Experimental design

```{r design}
data.frame(
  Group = c("TP53", "E6", "EmptyVector", "ATRX_FL", "ATRX_IFF"),
  Description = c("TP53 KO", "TP53 KO + ATRX KO",
                  "E6 + empty EGFP/mCherry vector",
                  "E6 + vector w/ ATRX full length (rescue)",
                  "E6 + vector w/ ATRX in-frame fusion (semi-rescue)"),
  `Parental clone` = c("TP53", "E6", "E6", "E6", "E6"),
  Vector = c("none", "none", "empty", "ATRX-FL", "ATRX-IFF"),
  n = c(3, 3, 3, 3, 3), check.names = FALSE
) %>% kable()
```

> Every group is a single clone, and each
> construct is a single transduced pool. The TP53 line and the E6 lineage differ
> by clonal background as well as by ATRX genotype, and those two cannot be
> separated in this design. Comparisons **within** the E6 lineage
> (`ATRX_FL` / `ATRX_IFF` / `EmptyVector`) are internally controlled and are the
> ones to trust; comparisons **against TP53** carry the clonal difference and are
> shown for completeness only.

---

# Quality control

## Library-level metrics

```{r qc-lib, fig.height = 6}
if (is.null(qc)) missing_note("design_qc", "design_qc_summary.csv") else {
  # "Reads assigned" alone has no denominator - a low bar could mean a small
  # library or a poor assignment rate. featureCounts' own total is recovered
  # from assigned_reads and pct_assigned so both are on the page.
  d <- qc %>% mutate(condition = cond_factor(condition),
                     sample = factor(sample, levels = meta$sample),
                     total_reads = if ("pct_assigned" %in% names(qc))
                       assigned_reads / (pct_assigned / 100) else NA_real_)
  lvls <- c("Total reads (featureCounts input)", "Reads assigned to genes",
            "% reads assigned", "Genes detected")
  print(
    d %>% select(sample, condition, total_reads, assigned_reads,
                 pct_assigned, genes_detected) %>%
      pivot_longer(c(total_reads, assigned_reads, pct_assigned, genes_detected)) %>%
      mutate(name = recode(name,
                           total_reads = "Total reads (featureCounts input)",
                           assigned_reads = "Reads assigned to genes",
                           pct_assigned = "% reads assigned",
                           genes_detected = "Genes detected"),
             name = factor(name, levels = lvls)) %>%
      filter(!is.na(value)) %>%
      ggplot(aes(sample, value, fill = condition)) +
      geom_col() +
      facet_wrap(~ name, scales = "free_y", ncol = 1) +
      scale_fill_manual(values = PAL) +
      scale_y_continuous(labels = function(x)
        format(x, big.mark = ",", scientific = FALSE, trim = TRUE)) +
      labs(x = NULL, y = NULL, fill = "Condition") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

> No group-linked shift in either depth, complexity or assignment rate. The libraries
> are comparable.

## ATRX and vector markers

Three ATRX features are quantified separately:

- **`ATRX`** — the endogenous locus in hg38, including its 5'/3'UTRs and introns.
- **`ATRX_FL`** — vector-borne full-length ATRX cDNA (no UTRs, no introns), fused
  N-terminally to a Twin-Strep-tag + 10xHis tag.
- **`ATRX_IFF`** — vector-borne in-frame-fusion ATRX cDNA carrying the *same*
  N-terminal tag, with exons 4–10 deleted (exon 3 joined directly to exon 10).

### What the figures actually count

Both constructs are cDNA copies of a gene already present in hg38, so most of
their sequence exists in **two or three copies** in the reference (vector
contigs + chrX). STAR flags such alignments as multi-mapping (`NH` ≥ 2,
MAPQ ≤ 3) and featureCounts is run with `-Q 10` and without `-M`, so these
fragments are **discarded, not distributed**. Only fragments placeable at
exactly one locus are counted:

- **`ATRX`** — fragments in the 5'/3'UTR (absent from the constructs), and
  fragments straddling an exon–intron boundary (introns are absent from the
  constructs). Fragments lying entirely within an intron are assigned to no
  gene; fragments lying entirely within coding exons are ambiguous with the
  constructs and discarded. `ATRX` therefore reports roughly the UTR-plus-
  boundary fraction of endogenous ATRX (~30% of the transcript), not its total
  expression — and it does so in *every* sample, since both vector contigs are
  in the reference regardless of which construct a sample carries.
- **`ATRX_FL`** — read *pairs* with one mate in the tag (absent from hg38) and
  the other mate extending past CDS position 133 into exons 4–10 (absent from
  `ATRX_IFF`). Neither mate is unique on its own; only the pair is.
- **`ATRX_IFF`** — fragments spanning the exon3–exon11 fusion breakpoint, which
  exists neither in hg38 nor in `ATRX_FL`.

> **Bar heights are not comparable between features.** The uniquely mappable
> window is ~100 nt wide for `ATRX_FL` and ~250–300 nt for `ATRX_IFF`, so
> `ATRX_FL` yields roughly 3× fewer counts than `ATRX_IFF` at equal expression.
> The FL/IFF difference in these plots is a mappability artefact, not
> differential transgene expression. Comparisons of the *same* feature across
> samples remain valid; comparisons *between* features do not.


```{r qc-markers, fig.height = 8}
mk <- gene_long(grab("design_qc", "marker_expression.csv"), meta)
if (is.null(mk)) missing_note("design_qc", "marker_expression.csv") else {
  print(
    ggplot(mk, aes(sample, value + 1, fill = condition)) +
      geom_col() +
      facet_wrap(~ gene_id, scales = "free_y", ncol = 1) +
      scale_y_log10() + scale_fill_manual(values = PAL) +
      labs(title = "Reporter and transgene expression",
           subtitle = "normalised counts + 1, log10; these features are excluded from the size factors",
           x = NULL, y = "normalised count + 1", fill = "Condition") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

```{r qc-atrx-table}
mk2 <- grab("design_qc", "marker_expression.csv")
if (!is.null(mk2) && "ATRX" %in% mk2$gene_id && !is.null(meta)) {
  gene_long(mk2, meta) %>% filter(gene_id == "ATRX") %>%
    group_by(condition) %>%
    summarise(mean = round(mean(value)), min = round(min(value)),
              max = round(max(value)), .groups = "drop") %>%
    mutate(`% of TP53` = round(100 * mean / mean[condition == "TP53"]),
           `fold spread` = sprintf("%.1fx", max / pmax(min, 1))) %>%
    kable(caption = "Endogenous ATRX locus, normalised counts")
}
```

> EGFP/mCherry are present only in the transduced groups; each ATRX construct is
> detected only in its own group. *(Add your read of the ATRX levels: is the
> rescue at a physiological dose? Is one construct variable across replicates?)*

## Sample–sample correlation

```{r qc-cor, fig.height = 6}
cm <- grab("design_qc", "sample_correlation.csv")
if (is.null(cm)) missing_note("design_qc", "sample_correlation.csv") else {
  m <- as.matrix(cm[, -1]); rownames(m) <- cm$sample
  ord <- meta$sample[meta$sample %in% rownames(m)]
  m <- m[ord, ord]
  print(
    as.data.frame(as.table(m)) %>%
      setNames(c("x", "y", "rho")) %>%
      mutate(x = factor(x, ord), y = factor(y, rev(ord))) %>%
      ggplot(aes(x, y, fill = rho)) +
      geom_tile(colour = "white", linewidth = 0.3) +
      scale_fill_gradient(low = "white", high = "#2166AC") +
      coord_fixed() +
      labs(title = "Spearman correlation (VST)", x = NULL, y = NULL) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            panel.grid = element_blank())
  )
}
```

> Replicates should form tight blocks on the diagonal. There is a very strong clonal effect between TP53 and E6 samples, that almost obfuscate the replicates correlation.

---

# Global structure (PCA)

```{r pca, fig.height = 5, results = "asis"}
# One view only: the E6-derived PCA has its own section at the end of the report.
# The ATRX features are removed here as well as the reporters - ATRX is the
# variable under study, so leaving it in makes the PCA partly re-plot the design
# rather than its downstream consequences.
pca_views <- list(
  c("all", "no_vector_no_transgene",
    "All samples, reporter and ATRX features removed")
)
for (v in pca_views) {
  co <- grab("pca", v[1], paste0(v[2], "_coords.csv"))
  ve <- grab("pca", v[1], paste0(v[2], "_variance_explained.csv"))
  cat("\n\n##", v[3], "\n\n")
  if (is.null(co)) { missing_note("pca", v[1], paste0(v[2], "_coords.csv")); next }
  pv <- if (!is.null(ve)) setNames(ve$percent_variance, ve$PC) else c(PC1 = NA, PC2 = NA)
  p <- co %>% mutate(condition = cond_factor(condition)) %>%
    ggplot(aes(PC1, PC2, colour = condition, label = sample)) +
    geom_point(size = 3) +
    ggrepel::geom_text_repel(size = 2.6, show.legend = FALSE, max.overlaps = 20) +
    scale_colour_manual(values = PAL) +
    labs(title = v[3],
         x = sprintf("PC1: %.1f%% variance", pv[["PC1"]]),
         y = sprintf("PC2: %.1f%% variance", pv[["PC2"]]),
         colour = "Condition")
  print(p)
}
```

> PC1 (78%) of the full PCA separates TP53 from everything E6-derived with :
> heavy clonal background, as seen with the sample-sample variance, confounded with the ATRX genotype and not correctable.
> Restricting to the E6 lineage lets the construct effects use PC1/PC2.

---

# How TP53 differs from the E6 lineage

Before narrowing to the E6 samples only, this is the size of the difference we are
setting aside. `TP53` is a different clone from everything E6-derived, and that
difference is confounded with the ATRX genotype.

```{r tp53-vs-e6, fig.height = 5}
r_tp53 <- de_results("deseq2", "TP53_vs_E6", "results.csv")
if (is.null(r_tp53)) missing_note("deseq2", "TP53_vs_E6", "results.csv") else {
  print(volcano(r_tp53, "TP53_vs_E6", num = "E6", den = "TP53"))
}
```

```{r tp53-scale}
if (!is.null(r_tp53)) {
  n <- de_counts(r_tp53)
  cat(sprintf(paste("\n> **%s genes** change between TP53 and E6 at padj < %s and",
                    "|log2FC| >= %s (%s up in E6, %s down) - roughly %.0f%% of the",
                    "genes tested. That is the clone plus the ATRX knockout",
                    "together, and it dwarfs everything that follows.\n\n"),
              format(sum(n), big.mark = ","), params$padj, params$lfc,
              format(n["up"], big.mark = ","), format(n["down"], big.mark = ","),
              100 * sum(n) / sum(!is.na(r_tp53$padj))))
}
```

```{r tp53-top}
if (!is.null(r_tp53)) kable(top_genes(r_tp53), caption = "TP53_vs_E6: top genes")
```

> With 18% of the genes teste being significantly differently expressed between TP53 and E6, 
> we can see thet either the clonal drift or the ATRX knockout led to huge shifts in expression.
> Our current set up does not ena

---

# Differential expression

## `r params$headline_contrast`

Both groups share the clone and carry a vector, so the clonal and transduction
effects cancel. This is the cleanest contrast in the dataset.

`r sprintf("The vector and transgene features (%s) are excluded from every differential-expression figure and table in this report. They are constructs rather than biology, and in any contrast crossing a construct boundary they would otherwise top the list.", paste(params$exclude_genes, collapse = ", "))`

Genes of interest (`r paste(params$highlight_genes, collapse = ", ")`) are
labelled in green on the volcano below.

```{r de-headline, fig.height = 5}
hc  <- params$headline_contrast
res <- de_results("deseq2", hc, "results.csv")
if (is.null(res)) missing_note("deseq2", hc, "results.csv") else {
  # EDIT the numerator/denominator if you change the headline contrast.
  print(volcano(res, hc, num = "ATRX_IFF", den = "ATRX_FL",
                highlight = params$highlight_genes))
}
```

```{r de-headline-top}
if (!is.null(res)) {
  kable(top_genes(res),
        caption = paste("Top 10 up and 10 down, ranked by -log10(padj) x |log2FC|",
                        "among genes passing both thresholds"))
}
```

> HOXB genes are overexpressed in `ATRX_IFF`. They are canonical targets of Polycomb Repressive Complex 2 (PRC2). 
> `ATRX_IFF` lacks its EZH2-interacting domain (PRC2 member) interacting domain, so one hypothesis would be
>  that `ATRX_IFF` fails recruiting PRC2 at the HOXB locus, therefore not repressing its expression.


## Genes of interest (HOXB and IGF1BP2)

The point of this panel is the **pattern across groups**, not any single gene.

```{r goi-load, include = FALSE}
goi <- gene_long(grab("design_qc", "goi_expression.csv"), meta)
```

```{r goi-means, fig.height = 4}
if (is.null(goi)) missing_note("design_qc", "goi_expression.csv") else {
  print(
    goi %>% group_by(gene_id, condition) %>%
      summarise(mean = mean(value), sd = sd(value), .groups = "drop") %>%
      ggplot(aes(condition, mean, fill = condition)) +
      geom_col() +
      geom_errorbar(aes(ymin = pmax(mean - sd, 0), ymax = mean + sd), width = 0.25,
                    linewidth = 0.3) +
      facet_wrap(~ gene_id, scales = "free_y", nrow = 1) +
      scale_fill_manual(values = PAL, guide = "none") +
      labs(title = "Group means (± SD)", x = NULL, y = "normalised count") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

> Genes significantly enriched in `ATRX_IFF` compared to `ATRX_FL` are also highly expressed 
> in TP53. This is counter-intuitive, as we would expect `TP53` cells to be more similar 
> to `ATRX_FL` (full rescue) than `ATRX_FF` (partial rescue via truncated protein)

## Remaining contrasts

```{r de-others, fig.height = 4.5, results = "asis"}
# EDIT: add or reorder. Each entry is c(contrast_dir, numerator, denominator).
contrasts <- list(
  c("EmptyVector_vs_ATRX_FL",  "ATRX_FL",     "EmptyVector"),
  c("EmptyVector_vs_ATRX_IFF", "ATRX_IFF",    "EmptyVector"),
  c("E6_vs_EmptyVector",       "EmptyVector", "E6"),
  # TP53_vs_E6 has its own section above.
  c("TP53_vs_ATRX_FL",         "ATRX_FL",     "TP53"),
  c("TP53_vs_ATRX_IFF",        "ATRX_IFF",    "TP53")
)
tally <- list()
for (c3 in contrasts) {
  r <- de_results("deseq2", c3[1], "results.csv")
  cat("\n\n###", c3[1], "\n\n")
  if (is.null(r)) { missing_note("deseq2", c3[1], "results.csv"); next }
  tally[[c3[1]]] <- de_counts(r)
  print(volcano(r, c3[1], num = c3[2], den = c3[3]))
}
```

```{r de-tally}
if (length(tally)) {
  do.call(rbind, tally) %>% as.data.frame() %>%
    tibble::rownames_to_column("contrast") %>%
    mutate(total = up + down) %>%
    kable(caption = sprintf("Genes with padj < %s and |log2FC| >= %s",
                            params$padj, params$lfc))
}
```

---

# Pathway enrichment (GSEA)

```{r gsea, fig.height = 6, results = "asis"}
gsea_dot <- function(df, title, top = 12) {
  df <- df %>% filter(!is.na(padj)) %>% arrange(desc(abs(NES)))
  df <- bind_rows(head(filter(df, NES > 0), top), head(filter(df, NES < 0), top))
  if (!nrow(df)) return(invisible(NULL))
  num <- df$numerator[1]; den <- df$denominator[1]
  up   <- sprintf("NES > 0: pathway genes HIGHER in %s", num)
  down <- sprintf("NES < 0: pathway genes HIGHER in %s", den)
  df <- df %>% mutate(
    pathway = gsub("^HALLMARK_", "", pathway),
    pathway = factor(pathway, levels = pathway[order(df$NES)]),
    side = factor(ifelse(NES > 0, up, down), levels = c(up, down)))
  print(
    ggplot(df, aes(NES, pathway, colour = padj, size = size)) +
      geom_point() +
      geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3) +
      facet_grid(side ~ ., scales = "free_y", space = "free_y") +
      scale_colour_gradient(low = "#B2182B", high = "grey75", limits = c(0, 0.25)) +
      labs(title = title,
           subtitle = sprintf("enrichment of the %s / %s ranking", num, den),
           y = NULL, colour = "padj", size = "Set size") +
      theme(strip.text.y = element_text(angle = 0, size = 8))
  )
}
# EDIT: which contrasts to show enrichment for.
for (cn in c("ATRX_FL_vs_ATRX_IFF", "EmptyVector_vs_ATRX_FL",
             "EmptyVector_vs_ATRX_IFF")) {
  g <- grab("gsea", cn, "H_results.csv")
  cat("\n\n##", cn, "\n\n")
  if (is.null(g)) { missing_note("gsea", cn, "H_results.csv"); next }
  gsea_dot(g, paste("Hallmark:", cn))
}
```

> Hallmark proliferation sets (E2F, G2M, MYC) are enriched in `ATRX_IFF` compared to `ATRX_FL`, and
> the inflammatory/interferon sets are enriched in `ATRX_FL`. 

---

# Rescue assessment

Whether the rescue worked is not answerable by asking if `ATRX_FL` clusters
with `TP53` — different clone, different transduction status. Instead: does the
intervention push the knockout-responsive genes back the other way? Both
contrasts used here sit inside a single genetic background, so the clonal
difference never enters.

## Signature reversal

```{r rescue-stats}
srs <- list.files(file.path(OUT, "signature_reversal"), pattern = "reversal_stats.csv",
                  recursive = TRUE, full.names = TRUE)
if (!length(srs)) missing_note("signature_reversal", "*", "reversal_stats.csv") else {
  # bind_rows, not rbind: stats files written by different versions of the script
  # can have different column sets, and rbind would abort the whole render.
  tab <- bind_rows(lapply(srs, read.csv)) %>%
    select(any_of(c("comparison", "n_signature_genes", "pct_reversed",
                    "null_pct_reversed", "excess_over_null", "perm_p",
                    "spearman_rho", "interpretable")))
  if ("excess_over_null" %in% names(tab)) tab <- arrange(tab, desc(excess_over_null))
  stale <- setdiff(c("null_pct_reversed", "excess_over_null", "perm_p"), names(tab))
  if (length(stale)) {
    cat("\n> **Note:** some `reversal_stats.csv` files predate the permutation",
        "null and lack", paste0("`", stale, "`", collapse = ", "),
        "- re-run those comparisons.\n\n")
  }
  kable(tab, caption = paste("Reversal of the knockout signature. Read the excess",
                             "over the permutation null, not the raw percentage."))
}
```

```{r rescue-scatter, fig.height = 6, results = "asis"}
# EDIT: which comparison to draw in full. The negative control pairs the vector-
# transduction signature (E6 vs EmptyVector) against the FL-vs-IFF difference -
# two groups that share no sample with the signature and have no mechanistic
# reason to reverse it. It calibrates the two rescue scores above: if it scores
# near the permutation null, the FL/IFF reversal is ATRX-specific; if it scores
# like ATRX_FL does, the metric is picking up generic structure instead.
for (nm in c("ATRX_FL_reverses_KO_signature_lfc1",
             "ATRX_IFF_reverses_KO_signature_lfc1",
             "NEGATIVE_CONTROL_vector_signature_vs_FL_IFF")) {
  g <- grab("signature_reversal", nm, "reversal_genes.csv")
  heading <- if (nm == "NEGATIVE_CONTROL_vector_signature_vs_FL_IFF")
    paste(nm, "(negative control)") else nm
  cat("\n\n###", heading, "\n\n")
  if (is.null(g)) { missing_note("signature_reversal", nm, "reversal_genes.csv"); next }
  lim <- max(abs(c(g$lfc_sig, g$lfc_resp)), na.rm = TRUE)
  st  <- grab("signature_reversal", nm, "reversal_stats.csv")
  # Stats go inside the panel: a long subtitle is silently clipped once the
  # figure is embedded or scaled, and these numbers are the point of the plot.
  box <- if (!is.null(st)) paste(
    sprintf("slope     %+.2f", st$tls_slope[1]),
    sprintf("R2        %.3f", if ("r_squared" %in% names(st))
              st$r_squared[1] else st$pearson_r[1]^2),
    sprintf("rho       %+.2f", st$spearman_rho[1]),
    sprintf("reversed  %.0f%% (null %.0f%%)", st$pct_reversed[1],
            st$null_pct_reversed[1]),
    sprintf("perm p    %s", format(signif(st$perm_p[1], 2))), sep = "\n") else NULL
  is_control <- nm == "NEGATIVE_CONTROL_vector_signature_vs_FL_IFF"
  p <- ggplot(g, aes(lfc_sig, lfc_resp)) +
    annotate("rect", xmin = 0, xmax = lim, ymin = -lim, ymax = 0,
             fill = "#4DAF4A", alpha = .07) +
    annotate("rect", xmin = -lim, xmax = 0, ymin = 0, ymax = lim,
             fill = "#4DAF4A", alpha = .07) +
    geom_hline(yintercept = 0, colour = "grey60", linewidth = .3) +
    geom_vline(xintercept = 0, colour = "grey60", linewidth = .3) +
    geom_abline(slope = -1, linetype = "dashed", colour = "grey45") +
    geom_point(size = 1, alpha = .5, colour = "#2166AC") +
    coord_fixed(xlim = c(-lim, lim), ylim = c(-lim, lim)) +
    labs(title = heading,
         subtitle = paste0("shaded = reversal | dashed = complete reversal (slope -1)\n",
                           "red = fitted slope (total least squares)"),
         x = if (is_control) "log2FC transduction (E6 / EmptyVector)"
             else "log2FC knockout (the damage)",
         y = if (is_control) "log2FC FL vs IFF (unrelated comparison)"
             else "log2FC rescue (the repair)")
  # The fitted slope is the statistic being reported, so draw it rather than
  # leaving the reader to infer it from the box.
  if (!is.null(st) && "tls_slope" %in% names(st) && is.finite(st$tls_slope[1])) {
    p <- p + geom_abline(slope = st$tls_slope[1], intercept = 0,
                         colour = "#E41A1C", linewidth = .8)
  }
  if (!is.null(box)) {
    p <- p + annotate("label", x = lim, y = lim, hjust = 1, vjust = 1, label = box,
                      size = 2.9, family = "mono", lineheight = 1.15,
                      fill = grDevices::adjustcolor("white", alpha.f = .88),
                      label.size = .25, colour = "grey15")
  }
  print(p)
}
```

> If a gene was pushed up by the knockout it should be pushed back down when reintroducing ATRX: points in the shaded
> quadrants, along the dashed line. Slope −1 is a complete rescue, 0 is no
> response. We see a slightly higher recovery in `ATRX_FL` compared to `ATRX_IFF`,
> however, both are very close to zero.

## Interaction contrasts

`(rescue / EmptyVector) − (TP53 / E6)`: how far the rescue moved each gene minus
how far it needed to move. Zero is a complete rescue; negative is a shortfall.

```{r rescue-interaction, fig.height = 4.5, results = "asis"}
for (nm in c("ATRX_FL_rescue_shortfall", "ATRX_IFF_rescue_shortfall")) {
  r <- de_results("deseq2_interaction", nm, "results.csv")
  cat("\n\n###", nm, "\n\n")
  if (is.null(r)) { missing_note("deseq2_interaction", nm, "results.csv"); next }
  print(volcano(r, nm, num = "overshoot", den = "shortfall"))
}
```

---

# The E6 lineage alone

With `TP53` excluded, every remaining group shares one clone and one
transduction step. Differences here are attributable to the construct — the two
effects we could not separate above are now held constant by design rather than
corrected for.

## PCA, E6-derived samples only

```{r e6-pca, fig.height = 5}
co <- grab("pca", "E6_derived", "no_vector_no_transgene_coords.csv")
ve <- grab("pca", "E6_derived", "no_vector_no_transgene_variance_explained.csv")
if (is.null(co)) {
  missing_note("pca", "E6_derived", "no_vector_no_transgene_coords.csv")
} else {
  pv <- if (!is.null(ve)) setNames(ve$percent_variance, ve$PC) else c(PC1=NA, PC2=NA)
  print(
    co %>% mutate(condition = cond_factor(condition)) %>%
      ggplot(aes(PC1, PC2, colour = condition, label = sample)) +
      geom_point(size = 3) +
      ggrepel::geom_text_repel(size = 2.6, show.legend = FALSE, max.overlaps = 20) +
      scale_colour_manual(values = PAL) +
      labs(title = "E6-derived samples, reporter and ATRX features removed",
           subtitle = "clone and transduction held constant; the construct is the only intended difference",
           x = sprintf("PC1: %.1f%% variance", pv[["PC1"]]),
           y = sprintf("PC2: %.1f%% variance", pv[["PC2"]]),
           colour = "Condition")
  )
}
```

## Sample–sample correlation, E6-derived only

Pairwise correlations are unchanged by dropping TP53 — but the colour scale
rescales to the range that remains, so differences compressed by the clonal
outlier become visible.

```{r e6-cor, fig.height = 5.5}
if (is.null(cm)) missing_note("design_qc", "sample_correlation.csv") else {
  m <- as.matrix(cm[, -1]); rownames(m) <- cm$sample
  keep <- meta$sample[meta$condition != "TP53" & meta$sample %in% rownames(m)]
  if (length(keep) < 3) {
    cat("\n> Fewer than 3 non-TP53 samples — nothing to plot.\n\n")
  } else {
    m <- m[keep, keep]
    print(
      as.data.frame(as.table(m)) %>%
        setNames(c("x", "y", "rho")) %>%
        mutate(x = factor(x, keep), y = factor(y, rev(keep))) %>%
        ggplot(aes(x, y, fill = rho)) +
        geom_tile(colour = "white", linewidth = 0.3) +
        geom_text(aes(label = sprintf("%.3f", rho)), size = 2.1, colour = "grey20") +
        scale_fill_gradient(low = "white", high = "#2166AC") +
        coord_fixed() +
        labs(title = "Spearman correlation (VST), TP53 excluded",
             subtitle = "scale rescaled to the E6 lineage",
             x = NULL, y = NULL) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
              panel.grid = element_blank())
    )
  }
}
```

## The three construct comparisons

```{r e6-three, fig.height = 4.5, results = "asis"}
trio <- list(
  c("EmptyVector_vs_ATRX_FL",  "ATRX_FL",  "EmptyVector",
    "Full-length rescue against its own transduced control"),
  c("EmptyVector_vs_ATRX_IFF", "ATRX_IFF", "EmptyVector",
    "In-frame fusion against the same control"),
  c("ATRX_FL_vs_ATRX_IFF",     "ATRX_IFF", "ATRX_FL",
    "The two constructs against each other")
)
trio_tally <- list()
for (t4 in trio) {
  r <- de_results("deseq2", t4[1], "results.csv")
  cat("\n\n###", t4[4], "\n\n")
  if (is.null(r)) { missing_note("deseq2", t4[1], "results.csv"); next }
  trio_tally[[t4[1]]] <- de_counts(r)
  print(volcano(r, t4[1], num = t4[2], den = t4[3]))
  cat("\n\n")
  print(kable(top_genes(r, n = 8)))
  cat("\n\n")
}
```

```{r e6-trio-tally}
if (length(trio_tally)) {
  do.call(rbind, trio_tally) %>% as.data.frame() %>%
    tibble::rownames_to_column("contrast") %>% mutate(total = up + down) %>%
    kable(caption = "Differentially expressed genes within the E6 lineage")
}
```

## Genes of interest, E6 lineage only

```{r e6-goi, fig.height = 4}
if (!is.null(goi)) {
  print(
    goi %>% filter(condition != "TP53") %>%
      group_by(gene_id, condition) %>%
      summarise(mean = mean(value), sd = sd(value), .groups = "drop") %>%
      ggplot(aes(condition, mean, fill = condition)) +
      geom_col() +
      geom_errorbar(aes(ymin = pmax(mean - sd, 0), ymax = mean + sd),
                    width = .25, linewidth = .3) +
      facet_wrap(~ gene_id, scales = "free_y", nrow = 1) +
      scale_fill_manual(values = PAL, guide = "none") +
      labs(title = "Genes of interest, TP53 excluded", x = NULL,
           y = "normalised count") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

> Here, we got more coherent results in both HOXB genes:
> taking E6 as baseline, we can see that HOXB expression is always lower in `ATRX_FL` 
> than `E6`, while it is much higher in `ATRX_IFF`. Interestingly, this could mean that `ATRX_IFF` has an acitve role in disturbing PRC2
> 

---

# Summary

TBD

```{r session, results = "asis"}
cat("\n<details><summary>Session info</summary>\n\n```\n")
cat(capture.output(sessionInfo()), sep = "\n")
cat("\n```\n</details>\n")
```
