---
title: "Epicode: ATRX rescue in an ATRX-KO neuroblastoma line"
subtitle: "Bulk RNA-seq analysis"
author: "Arthur Dondi"
date: "`r format(Sys.Date(), '%d %B %Y')`"
params:
  # Point this at User.output_dir from your config, or pass it at render time.
  outdir: "/nobackup/lab_taschner-mandl/arthurdondi/projects/epicode"
  padj: 0.05
  lfc: 1.0
  # Order groups appear in every figure. Edit if you add conditions.
  condition_levels: !r c("TP53", "E6", "EmptyVector", "ATRX_FL", "ATRX_IFF")
  # The contrast the report leads with.
  headline_contrast: "ATRX_IFF_vs_ATRX_FL"
  # Vector and transgene features. They are rows in the count matrix because the
  # reference carries the vector contigs, but they are constructs rather than
  # biology, so they are dropped from every differential-expression figure and
  # table below. Removed here in the report only - the pipeline's own outputs
  # still contain them.
  exclude_genes: !r c("ATRX", "ATRX_FL", "ATRX_IFF", "EGFP", "mCherry")
  # Labelled on the headline volcano.
  highlight_genes: !r c("HOXB2", "HOXB4", "HOXB6", "HOXB7", "IGF2BP1")
  # Every volcano additionally labels its N most significant genes (by
  # -log10(padj) x |log2FC|, the same ranking as the top-genes tables), the way
  # the pipeline's own volcano.pdf labels every significant gene. The report
  # caps it at a fixed N instead, since some contrasts have thousands of
  # significant genes and labelling all of them is unreadable at report scale.
  volcano_label_top: 12
  # Optional proteomics overlay for the headline ATRX_IFF_vs_ATRX_FL volcano.
  # Mirrors Proteomics: in config_epicode.yaml. Point proteomics_xlsx at "" (or
  # a path that does not exist) to skip the panel gracefully.
  proteomics_xlsx: "/nobackup/lab_taschner-mandl/arthurdondi/projects/epicode/251128_limmaResult_IP_MRa_051_LFQ_Experiment1_V3_filtered-KK.xlsx"
  proteomics_sheet: "limma result filtered"
  proteomics_gene_column: "Gene"
  proteomics_fdr_column: "_FDR"
  proteomics_logfc_column: "_LOG2FC"
  proteomics_fdr_threshold: 0.05
  proteomics_lfc_threshold: 1.0
  # DESeq2_to_proteomics comparison for the headline contrast (see the DIRECTION
  # note in config_epicode.yaml: proteomics columns read {numerator}_vs_{denominator},
  # the same convention as the DESeq2 contrast name).
  proteomics_comparison: "S1744_vs_S1011"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    self_contained: true
    theme: flatly
    fig_width: 9
    fig_height: 5
    df_print: kable
    # Every chunk's code is included but folded behind a "Show" toggle, so the
    # report reads as figures and tables by default while staying fully
    # reproducible one click away. Requires echo = TRUE in the setup chunk below
    # - code_folding only folds code that would otherwise be printed.
    code_folding: hide
  pdf_document:
    toc: true
    number_sections: true
    fig_width: 8
    fig_height: 4.5
---

<!--
HOW TO RENDER
  Rscript -e 'rmarkdown::render("report/analysis_report.Rmd",
              params = list(outdir = "/path/to/output_dir"),
              output_file = "analysis_report.html")'

  Swap output_format = "pdf_document" for PDF (needs LaTeX).

HOW THIS WORKS
  Every figure is rebuilt here from the CSV files the pipeline already writes,
  rather than embedding the PDFs. That keeps the styling consistent, makes the
  HTML self-contained, and means you can edit any plot in this file.

  Every section degrades gracefully: if an output is missing the section says so
  and the report still renders. So you can run it before the pipeline finishes.

  Text in > blockquotes is placeholder interpretation - rewrite it.
-->

```{r setup, include = FALSE}
# echo = TRUE + code_folding: hide (in the YAML header) is the "Show" toggle
# per chunk - code is present in the HTML but folded away by default.
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      dpi = 150, fig.align = "center")

suppressPackageStartupMessages({
  library(ggplot2); library(dplyr); library(tidyr); library(knitr)
})

OUT <- params$outdir
LEV <- params$condition_levels

# Condition palette: the ggplot hue values, so figures match the pipeline PDFs.
PAL <- setNames(
  c("#E76BF3", "#00BF7D", "#00B0F6", "#F8766D", "#A3A500")[seq_along(LEV)], LEV)

theme_set(theme_bw(base_size = 11))
theme_update(panel.grid.minor = element_blank(),
             plot.title = element_text(face = "bold", size = 12),
             plot.subtitle = element_text(colour = "grey30", size = 9))

# ── Helpers ──────────────────────────────────────────────────────────────────
# Read a pipeline output, or return NULL and leave a visible note in the report.
grab <- function(...) {
  f <- file.path(OUT, ...)
  if (!file.exists(f)) return(NULL)
  read.csv(f, stringsAsFactors = FALSE, check.names = FALSE)
}
missing_note <- function(...) {
  cat("\n> **Not available.** `", file.path(...), "` was not found. ",
      "Run that part of the pipeline and re-render.\n\n", sep = "")
}
cond_factor <- function(x) factor(x, levels = LEV[LEV %in% unique(x)])

# Long-format helper for the wide per-gene tables (gene_id + one column per sample).
gene_long <- function(df, meta) {
  if (is.null(df) || nrow(df) == 0) return(NULL)
  df %>%
    pivot_longer(-gene_id, names_to = "sample", values_to = "value") %>%
    left_join(meta, by = "sample") %>%
    mutate(condition = cond_factor(condition),
           sample = factor(sample, levels = meta$sample))
}

# Sample sheet, used to colour everything by condition.
qc <- grab("design_qc", "design_qc_summary.csv")
meta <- if (!is.null(qc)) {
  qc %>% select(sample, condition) %>%
    arrange(match(condition, LEV)) %>%
    mutate(condition = cond_factor(condition))
} else NULL

# DESeq results with the vector/transgene features dropped. They are guaranteed
# to be the most significant genes in any contrast that crosses a construct
# boundary, and they say nothing about the cells' biology.
de_results <- function(...) {
  r <- grab(...)
  if (is.null(r)) return(NULL)
  n0 <- nrow(r)
  r  <- r[!tolower(r$gene_id) %in% tolower(params$exclude_genes), , drop = FALSE]
  attr(r, "n_dropped") <- n0 - nrow(r)
  r
}

volcano <- function(res, title, num, den, highlight = NULL,
                    label_top = params$volcano_label_top) {
  res <- res %>% filter(!is.na(padj)) %>%
    mutate(sig = padj < params$padj & abs(log2FoldChange) >= params$lfc)
  n <- c(up = sum(res$sig & res$log2FoldChange > 0),
         down = sum(res$sig & res$log2FoldChange < 0))
  p <- ggplot(res, aes(log2FoldChange, -log10(padj), colour = sig)) +
    geom_point(size = 0.8, alpha = 0.5) +
    geom_vline(xintercept = c(-params$lfc, params$lfc), linetype = "dashed",
               linewidth = 0.3) +
    geom_hline(yintercept = -log10(params$padj), linetype = "dashed",
               linewidth = 0.3) +
    scale_colour_manual(values = c("FALSE" = "grey70", "TRUE" = "#B2182B"),
                        guide = "none") +
    labs(
      title = title,
      # Spell the direction out on both axes: "up in X" is ambiguous unless the
      # reader already knows which group is the numerator.
      subtitle = sprintf(
        paste0("%s significant (padj < %s, |log2FC| >= %s)\n",
               "RIGHT, log2FC > 0: %s genes HIGHER in %s   |   ",
               "LEFT, log2FC < 0: %s genes HIGHER in %s"),
        format(sum(res$sig), big.mark = ","), params$padj, params$lfc,
        format(n["up"], big.mark = ","), num,
        format(n["down"], big.mark = ","), den),
      x = sprintf("log2FC  (%s / %s)   \u2014   positive = higher in %s", num, den, num),
      y = expression(-log[10] ~ "adjusted p")) +
    theme(plot.subtitle = element_text(lineheight = 1.2))
  # Label the most significant genes, ranked the same way as top_genes():
  # -log10(padj) x |log2FC|, so the labelled genes match the accompanying table.
  # The pipeline's own volcano.pdf labels every significant gene; capped here
  # because report-scale figures become unreadable once a contrast has more
  # than a few dozen significant genes (E6_vs_TP53 has thousands).
  hl_lower <- if (is.null(highlight)) character(0) else tolower(highlight)
  if (label_top > 0) {
    lbl <- res %>% filter(sig, !tolower(gene_id) %in% hl_lower) %>%
      mutate(score = -log10(padj) * abs(log2FoldChange)) %>%
      slice_max(score, n = label_top, with_ties = FALSE)
    if (nrow(lbl)) {
      p <- p + ggrepel::geom_text_repel(
        data = lbl, aes(label = gene_id), colour = "grey20", size = 2.4,
        max.overlaps = 30, segment.size = 0.2, segment.colour = "grey50",
        show.legend = FALSE)
    }
  }
  if (!is.null(highlight)) {
    h <- res %>% filter(tolower(gene_id) %in% hl_lower)
    if (nrow(h)) {
      # Drawn after the general labels so the highlighted genes and their
      # labels sit on top rather than being obscured by them.
      p <- p +
        geom_point(data = h, colour = "#1B7837", size = 2.4) +
        ggrepel::geom_text_repel(
          data = h, aes(label = gene_id), colour = "#1B7837", fontface = "bold",
          size = 3.1, min.segment.length = 0, box.padding = 0.6,
          segment.colour = "#1B7837", max.overlaps = Inf, show.legend = FALSE)
    }
    miss <- setdiff(hl_lower, tolower(res$gene_id))
    if (length(miss)) message("highlight not found: ", paste(miss, collapse = ", "))
  }
  p
}

# Ranking by padj alone floats high-count genes with tiny fold-changes to the
# top. This ranks significant genes by -log10(padj) * |log2FC| so both matter,
# and reports the two directions separately so one cannot crowd out the other.
top_genes <- function(res, n = 10) {
  res %>% filter(!is.na(padj), padj < params$padj,
                 abs(log2FoldChange) >= params$lfc) %>%
    mutate(direction = ifelse(log2FoldChange > 0, "up", "down"),
           score = -log10(padj) * abs(log2FoldChange)) %>%
    group_by(direction) %>% slice_max(score, n = n) %>% ungroup() %>%
    arrange(direction, desc(score)) %>%
    transmute(gene_id, direction, baseMean = round(baseMean),
              log2FC = round(log2FoldChange, 2), padj = signif(padj, 3),
              score = round(score, 1))
}

de_counts <- function(res) {
  r <- res %>% filter(!is.na(padj), padj < params$padj,
                      abs(log2FoldChange) >= params$lfc)
  c(up = sum(r$log2FoldChange > 0), down = sum(r$log2FoldChange < 0))
}

# Proteomics-significant genes for one comparison, wide-format limma Excel.
# Mirrors workflow/scripts/deseq2_proteomic.R. Returns NULL (and lets the
# caller show missing_note()) if the file, sheet or columns are not there.
load_proteomics_sig <- function() {
  if (!nchar(params$proteomics_xlsx) || !file.exists(params$proteomics_xlsx)) return(NULL)
  logfc_col <- paste0(params$proteomics_comparison, params$proteomics_logfc_column)
  fdr_col   <- paste0(params$proteomics_comparison, params$proteomics_fdr_column)
  tbl <- tryCatch(
    readxl::read_excel(params$proteomics_xlsx, sheet = params$proteomics_sheet, col_types = "text"),
    error = function(e) NULL)
  if (is.null(tbl) || !all(c(params$proteomics_gene_column, logfc_col, fdr_col) %in% colnames(tbl)))
    return(NULL)
  tbl %>%
    transmute(gene_id    = trimws(as.character(.data[[params$proteomics_gene_column]])),
              prot_fdr   = suppressWarnings(as.numeric(.data[[fdr_col]])),
              prot_logfc = suppressWarnings(as.numeric(.data[[logfc_col]]))) %>%
    filter(!is.na(gene_id), gene_id != "", !is.na(prot_fdr), !is.na(prot_logfc),
           prot_fdr <= params$proteomics_fdr_threshold,
           abs(prot_logfc) > params$proteomics_lfc_threshold) %>%
    arrange(prot_fdr, dplyr::desc(abs(prot_logfc))) %>%
    distinct(gene_id, .keep_all = TRUE) %>%
    transmute(gene_id, prot_direction = ifelse(prot_logfc > 0, "Up", "Down"))
}

# RNA volcano restricted to genes that are also proteomics-significant, coloured
# by RNA/protein direction agreement - same construction as the pipeline's own
# volcano_proteomic.pdf, redrawn here in the report's own style.
proteomic_volcano <- function(res, title, num, den, prot_sig) {
  res <- res %>% filter(!is.na(padj)) %>%
    mutate(rna_significant = padj < params$padj & abs(log2FoldChange) >= params$lfc,
           rna_direction = ifelse(log2FoldChange > 0, "Up", "Down"))
  df <- res %>% inner_join(prot_sig, by = "gene_id") %>%
    mutate(significance = dplyr::case_when(
      rna_significant & rna_direction == prot_direction ~ "Significant same direction",
      rna_significant & rna_direction != prot_direction ~ "Significant opposite direction",
      TRUE ~ "Not significant"),
      label = ifelse(significance != "Not significant", gene_id, NA_character_))
  if (!nrow(df)) return(invisible(NULL))
  ggplot(df, aes(log2FoldChange, -log10(padj), colour = significance)) +
    geom_point(alpha = 0.6, size = 1.2) +
    ggrepel::geom_text_repel(aes(label = label), size = 2.5, max.overlaps = Inf,
                             show.legend = FALSE) +
    scale_colour_manual(values = c("Significant same direction" = "#33A02C",
                                   "Significant opposite direction" = "#E31A1C",
                                   "Not significant" = "grey60")) +
    geom_vline(xintercept = c(-params$lfc, params$lfc), linetype = "dashed", linewidth = 0.3) +
    geom_hline(yintercept = -log10(params$padj), linetype = "dashed", linewidth = 0.3) +
    labs(
      title = title,
      subtitle = sprintf(
        paste0("Proteomics-significant genes only (FDR <= %s, |log2FC| > %s)\n",
               "RIGHT, log2FC > 0: higher in %s   |   LEFT, log2FC < 0: higher in %s\n",
               "green = RNA and protein agree   |   red = RNA and protein disagree"),
        params$proteomics_fdr_threshold, params$proteomics_lfc_threshold, num, den),
      x = sprintf("log2FC  (%s / %s)   —   positive = higher in %s", num, den, num),
      y = expression(-log[10] ~ "adjusted p"), colour = NULL) +
    theme(plot.subtitle = element_text(lineheight = 1.2, size = 8.5), legend.position = "bottom")
}
```

# Experimental design

```{r design}
data.frame(
  Group = c("TP53", "E6", "EmptyVector", "ATRX_FL", "ATRX_IFF"),
  Description = c("TP53 KO", "TP53 KO + ATRX KO",
                  "E6 + empty EGFP/mCherry vector",
                  "E6 + vector w/ ATRX full length (rescue)",
                  "E6 + vector w/ ATRX in-frame fusion (semi-rescue)"),
  `Parental clone` = c("TP53", "E6", "E6", "E6", "E6"),
  Vector = c("none", "none", "empty", "ATRX-FL", "ATRX-IFF"),
  n = c(3, 3, 3, 3, 3), check.names = FALSE
) %>% kable()
```

> Every group is a single clone, and each
> construct is a single transduced pool. The TP53 line and the E6 lineage differ
> by clonal background as well as by ATRX genotype, and those two cannot be
> separated in this design. Comparisons **within** the E6 lineage
> (`ATRX_FL` / `ATRX_IFF` / `EmptyVector`) are internally controlled and are the
> ones to trust; comparisons **against TP53** carry the clonal difference and are
> shown for completeness only.

---

# Quality control

## Library-level metrics

```{r qc-lib, fig.height = 6}
if (is.null(qc)) missing_note("design_qc", "design_qc_summary.csv") else {
  # "Reads assigned" alone has no denominator - a low bar could mean a small
  # library or a poor assignment rate. featureCounts' own total is recovered
  # from assigned_reads and pct_assigned so both are on the page.
  d <- qc %>% mutate(condition = cond_factor(condition),
                     sample = factor(sample, levels = meta$sample),
                     total_reads = if ("pct_assigned" %in% names(qc))
                       assigned_reads / (pct_assigned / 100) else NA_real_)
  lvls <- c("Total reads (featureCounts input)", "Reads assigned to genes",
            "% reads assigned", "Genes detected")
  print(
    d %>% select(sample, condition, total_reads, assigned_reads,
                 pct_assigned, genes_detected) %>%
      pivot_longer(c(total_reads, assigned_reads, pct_assigned, genes_detected)) %>%
      mutate(name = recode(name,
                           total_reads = "Total reads (featureCounts input)",
                           assigned_reads = "Reads assigned to genes",
                           pct_assigned = "% reads assigned",
                           genes_detected = "Genes detected"),
             name = factor(name, levels = lvls)) %>%
      filter(!is.na(value)) %>%
      ggplot(aes(sample, value, fill = condition)) +
      geom_col() +
      facet_wrap(~ name, scales = "free_y", ncol = 1) +
      scale_fill_manual(values = PAL) +
      scale_y_continuous(labels = function(x)
        format(x, big.mark = ",", scientific = FALSE, trim = TRUE)) +
      labs(x = NULL, y = NULL, fill = "Condition") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

> No group-linked shift in either depth, complexity or assignment rate. The libraries
> are comparable.

## ATRX and vector markers

ATRX_wt is the wild-type version, it includes introns, 3'UTR and 5'UTR. ATRX_FL
(7482 bp) and ATRX_IFF (3993 bp) include a Twin-Strep-tag and 10xHis tag, and are
cDNAs (no 3'/5'UTR, no introns).

ATRX_wt, ATRX_FL and ATRX_IFF only account for uniquely mappable reads: reads that
are unambiguously assigned to one of the 3 possible references. That is:

- Reads mapping to either ATRX 3'/5'UTR and reads overlapping exons and introns
  (reads only spanning introns are not counted) are reported as `ATRX_wt`.
- Read pairs with one mate on the Twin-Strep/10His tag and the other past the
  FL/IFF divergence point (nt 250) are reported as `ATRX_FL`.
- Read pairs spanning the fusion breakpoint are reported as `ATRX_IFF`.

Two further rows are counted from the same alignments but keep the multi-mapping
reads:

- `ATRX_Total`: all reads from any ATRX source — wild-type, FL and IFF pooled, tag
  included. Much higher than the sum of the three rows above, which discard
  multimapped reads. Counts scale with transcript length, so the shorter IFF
  construct is under-represented 1.87-fold.
- `Tag`: reads on the Twin-Strep/10His tag only. Most reliable `ATRX_IFF` vs
  `ATRX_FL` expression comparison.

```{r qc-markers, fig.height = 11}
mk <- gene_long(grab("design_qc", "marker_expression.csv"), meta)
if (is.null(mk)) missing_note("design_qc", "marker_expression.csv") else {
  # Keep the CSV's row order (ATRX_Total and Tag first) instead of letting
  # ggplot sort the facets alphabetically, and match the pipeline's ATRX_wt
  # label. The CSV itself keeps the real feature IDs.
  relabel <- function(x) ifelse(x == "ATRX", "ATRX_wt", x)
  mk$gene_id <- factor(relabel(mk$gene_id), levels = relabel(unique(mk$gene_id)))
  print(
    ggplot(mk, aes(sample, value, fill = condition)) +
      geom_col() +
      facet_wrap(~ gene_id, scales = "free_y", ncol = 1) +
      scale_fill_manual(values = PAL) +
      labs(title = "Reporter and transgene expression",
           x = NULL, y = "normalised count", fill = "Condition") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

```{r qc-atrx-table}
mk2 <- grab("design_qc", "marker_expression.csv")
if (!is.null(mk2) && "ATRX" %in% mk2$gene_id && !is.null(meta)) {
  gene_long(mk2, meta) %>% filter(gene_id == "ATRX") %>%
    group_by(condition) %>%
    summarise(mean = round(mean(value)), min = round(min(value)),
              max = round(max(value)), .groups = "drop") %>%
    mutate(`% of TP53` = round(100 * mean / mean[condition == "TP53"]),
           `fold spread` = sprintf("%.1fx", max / pmax(min, 1))) %>%
    kable(caption = "ATRX_wt (endogenous locus), normalised counts")
}
```

> `EGFP`/`mCherry` are present only in the transduced groups, although with large
> variations; each ATRX construct is detected only in its own group. `E6` seems to
> have a reduced expression compared to `TP53`, possibly because the CRISPR
> modification in ATRX introduced a premature stop codon leading to
> nonsense-mediated decay. `ATRX_wt` and `ATRX_Total` show very similar profiles,
> which is expected for `TP53` and `E6`; however for `ATRX_FL` and `ATRX_IFF` one
> would expect a bump in expression due to multimapped construct reads. Via the
> `Tag` expression, we see that `ATRX_IFF` is much more expressed than `ATRX_FL`.

## Sample–sample correlation

```{r qc-cor, fig.height = 6}
cm <- grab("design_qc", "sample_correlation.csv")
if (is.null(cm)) missing_note("design_qc", "sample_correlation.csv") else {
  m <- as.matrix(cm[, -1]); rownames(m) <- cm$sample
  ord <- meta$sample[meta$sample %in% rownames(m)]
  m <- m[ord, ord]
  print(
    as.data.frame(as.table(m)) %>%
      setNames(c("x", "y", "rho")) %>%
      mutate(x = factor(x, ord), y = factor(y, rev(ord))) %>%
      ggplot(aes(x, y, fill = rho)) +
      geom_tile(colour = "white", linewidth = 0.3) +
      scale_fill_gradient(low = "white", high = "#2166AC") +
      coord_fixed() +
      labs(title = "Spearman correlation (VST)", x = NULL, y = NULL) +
      theme(axis.text.x = element_text(angle = 45, hjust = 1),
            panel.grid = element_blank())
  )
}
```

> Replicates should form tight blocks on the diagonal. There is a very strong clonal effect between TP53 and E6 samples, that almost obfuscate the replicates correlation.

---

# Global structure (PCA)

```{r pca, fig.height = 5, results = "asis"}
# One view only: the E6-derived PCA has its own section at the end of the report.
# The ATRX features are removed here as well as the reporters - ATRX is the
# variable under study, so leaving it in makes the PCA partly re-plot the design
# rather than its downstream consequences.
pca_views <- list(
  c("all", "no_vector_no_transgene",
    "All samples, reporter and ATRX features removed")
)
for (v in pca_views) {
  co <- grab("pca", v[1], paste0(v[2], "_coords.csv"))
  ve <- grab("pca", v[1], paste0(v[2], "_variance_explained.csv"))
  cat("\n\n##", v[3], "\n\n")
  if (is.null(co)) { missing_note("pca", v[1], paste0(v[2], "_coords.csv")); next }
  pv <- if (!is.null(ve)) setNames(ve$percent_variance, ve$PC) else c(PC1 = NA, PC2 = NA)
  p <- co %>% mutate(condition = cond_factor(condition)) %>%
    ggplot(aes(PC1, PC2, colour = condition, label = sample)) +
    geom_point(size = 3) +
    ggrepel::geom_text_repel(size = 2.6, show.legend = FALSE, max.overlaps = 20) +
    scale_colour_manual(values = PAL) +
    labs(title = v[3],
         x = sprintf("PC1: %.1f%% variance", pv[["PC1"]]),
         y = sprintf("PC2: %.1f%% variance", pv[["PC2"]]),
         colour = "Condition")
  print(p)
}
```

> PC1 (78%) of the full PCA separates TP53 from everything E6-derived with :
> heavy clonal background, as seen with the sample-sample variance, confounded with the ATRX genotype and not correctable.

---

# How TP53 differs from the E6 lineage

Before narrowing to the E6 samples only, this is the size of the difference we are
setting aside. `TP53` is a different clone from everything E6-derived, and that
difference is confounded with the ATRX genotype.

```{r tp53-vs-e6, fig.height = 5}
r_tp53 <- de_results("deseq2", "E6_vs_TP53", "results.csv")
if (is.null(r_tp53)) missing_note("deseq2", "E6_vs_TP53", "results.csv") else {
  print(volcano(r_tp53, "E6_vs_TP53", num = "E6", den = "TP53"))
}
```

```{r tp53-scale}
if (!is.null(r_tp53)) {
  n <- de_counts(r_tp53)
  cat(sprintf(paste("\n> **%s genes** change between TP53 and E6 at padj < %s and",
                    "|log2FC| >= %s (%s up in E6, %s down) - roughly %.0f%% of the",
                    "genes tested. That is the clone plus the ATRX knockout",
                    "together, and it dwarfs everything that follows.\n\n"),
              format(sum(n), big.mark = ","), params$padj, params$lfc,
              format(n["up"], big.mark = ","), format(n["down"], big.mark = ","),
              100 * sum(n) / sum(!is.na(r_tp53$padj))))
}
```

```{r tp53-top}
if (!is.null(r_tp53)) kable(top_genes(r_tp53), caption = "E6_vs_TP53: top genes")
```

> With 18% of the genes teste being significantly differently expressed between TP53 and E6, 
> we can see that either the clonal drift or the ATRX knockout led to huge shifts in expression.
> Our current set up does not enable us to separate the two, clonal background and ATRX status vary together and are perfectly confounded


## Differential expression across contrasts

Gene counts only. The per-contrast volcanoes are in the pipeline output under
`deseq2/{contrast}/`; what matters here is the relative size of each effect.

`r sprintf("The vector and transgene features (%s) are excluded from every differential-expression figure and table in this report. They are constructs rather than biology, and in any contrast crossing a construct boundary they would otherwise top the list.", paste(params$exclude_genes, collapse = ", "))`

```{r de-tally}
# Self-contained: reads each contrast's results and counts, without drawing
# anything. The headline ATRX_FL vs ATRX_IFF comparison is included so the
# cleanest contrast in the design can be sized against the rest.
tally_contrasts <- c(
  params$headline_contrast,
  "ATRX_FL_vs_EmptyVector",
  "ATRX_IFF_vs_EmptyVector",
  "EmptyVector_vs_E6",
  "E6_vs_TP53",
  "ATRX_FL_vs_TP53",
  "ATRX_IFF_vs_TP53"
)
tally <- list()
for (cn in tally_contrasts) {
  r <- de_results("deseq2", cn, "results.csv")
  if (!is.null(r)) tally[[cn]] <- de_counts(r)
}
if (length(tally)) {
  do.call(rbind, tally) %>% as.data.frame() %>%
    tibble::rownames_to_column("contrast") %>%
    mutate(total = up + down) %>%
    kable(caption = sprintf("Genes with padj < %s and |log2FC| >= %s",
                            params$padj, params$lfc))
} else {
  missing_note("deseq2", "*", "results.csv")
}
```

---

# Rescue assessment

Whether the rescue worked is not answerable by asking if `ATRX_FL` clusters
with `TP53` — different clone, different transduction status. Instead: does the
intervention push the knockout-responsive genes back the other way? Both
contrasts used here sit inside a single genetic background, so the clonal
difference never enters.

## Signature reversal

```{r rescue-stats}
srs <- list.files(file.path(OUT, "signature_reversal"), pattern = "reversal_stats.csv",
                  recursive = TRUE, full.names = TRUE)
if (!length(srs)) missing_note("signature_reversal", "*", "reversal_stats.csv") else {
  # bind_rows, not rbind: stats files written by different versions of the script
  # can have different column sets, and rbind would abort the whole render.
  tab <- bind_rows(lapply(srs, read.csv)) %>%
    select(any_of(c("comparison", "n_signature_genes", "pct_reversed",
                    "null_pct_reversed", "excess_over_null", "perm_p",
                    "spearman_rho", "interpretable")))
  if ("excess_over_null" %in% names(tab)) tab <- arrange(tab, desc(excess_over_null))
  stale <- setdiff(c("null_pct_reversed", "excess_over_null", "perm_p"), names(tab))
  if (length(stale)) {
    cat("\n> **Note:** some `reversal_stats.csv` files predate the permutation",
        "null and lack", paste0("`", stale, "`", collapse = ", "),
        "- re-run those comparisons.\n\n")
  }
  kable(tab, caption = paste("Reversal of the knockout signature. Read the excess",
                             "over the permutation null, not the raw percentage."))
}
```

```{r rescue-scatter, fig.height = 6, results = "asis"}
# EDIT: which comparison to draw in full. The negative control pairs the vector-
# transduction signature (E6 vs EmptyVector) against the FL-vs-IFF difference -
# two groups that share no sample with the signature and have no mechanistic
# reason to reverse it. It calibrates the two rescue scores above: if it scores
# near the permutation null, the FL/IFF reversal is ATRX-specific; if it scores
# like ATRX_FL does, the metric is picking up generic structure instead.
for (nm in c("ATRX_FL_reverses_KO_signature_lfc1",
             "ATRX_IFF_reverses_KO_signature_lfc1",
             "NEGATIVE_CONTROL_vector_signature_vs_FL_IFF")) {
  g <- grab("signature_reversal", nm, "reversal_genes.csv")
  heading <- if (nm == "NEGATIVE_CONTROL_vector_signature_vs_FL_IFF")
    paste(nm, "(negative control)") else nm
  cat("\n\n###", heading, "\n\n")
  if (is.null(g)) { missing_note("signature_reversal", nm, "reversal_genes.csv"); next }
  # Report-only: drop the vector/transgene features, same as every DE table and
  # volcano above (de_results()). The negative control pairs EmptyVector_vs_E6
  # as its signature, so EGFP/mCherry would otherwise dominate it.
  g <- g[!tolower(g$gene_id) %in% tolower(params$exclude_genes), , drop = FALSE]
  if (!nrow(g)) { cat("\n> No genes left after excluding vector/transgene features.\n\n"); next }
  lim <- max(abs(c(g$lfc_sig, g$lfc_resp)), na.rm = TRUE)
  st  <- grab("signature_reversal", nm, "reversal_stats.csv")
  # Stats go inside the panel: a long subtitle is silently clipped once the
  # figure is embedded or scaled, and these numbers are the point of the plot.
  box <- if (!is.null(st)) paste(
    sprintf("slope     %+.2f", st$tls_slope[1]),
    sprintf("R2        %.3f", if ("r_squared" %in% names(st))
              st$r_squared[1] else st$pearson_r[1]^2),
    sprintf("rho       %+.2f", st$spearman_rho[1]),
    sprintf("reversed  %.0f%% (null %.0f%%)", st$pct_reversed[1],
            st$null_pct_reversed[1]),
    sprintf("perm p    %s", format(signif(st$perm_p[1], 2))), sep = "\n") else NULL
  is_control <- nm == "NEGATIVE_CONTROL_vector_signature_vs_FL_IFF"
  # Name the actual contrast on each axis, e.g. "E6 / TP53", not a generic
  # "the damage" - matching the pipeline's own reversal_scatter.pdf. Read from
  # reversal_stats.csv (signature_contrast / response_contrast) rather than
  # hardcoding, since that is the one place the exact label is guaranteed to
  # match what the stats box above was computed from.
  sig_label  <- if (!is.null(st) && "signature_contrast" %in% names(st))
    st$signature_contrast[1] else "signature"
  resp_label <- if (!is.null(st) && "response_contrast" %in% names(st))
    st$response_contrast[1] else "response"
  p <- ggplot(g, aes(lfc_sig, lfc_resp)) +
    annotate("rect", xmin = 0, xmax = lim, ymin = -lim, ymax = 0,
             fill = "#4DAF4A", alpha = .07) +
    annotate("rect", xmin = -lim, xmax = 0, ymin = 0, ymax = lim,
             fill = "#4DAF4A", alpha = .07) +
    geom_hline(yintercept = 0, colour = "grey60", linewidth = .3) +
    geom_vline(xintercept = 0, colour = "grey60", linewidth = .3) +
    geom_abline(slope = -1, linetype = "dashed", colour = "grey45") +
    geom_point(size = 1, alpha = .5, colour = "#2166AC") +
    coord_fixed(xlim = c(-lim, lim), ylim = c(-lim, lim)) +
    labs(title = heading,
         subtitle = paste0("shaded = reversal | dashed = complete reversal (slope -1)\n",
                           "red = fitted slope (total least squares)"),
         x = sprintf("log2FC  %s%s", sig_label,
                     if (is_control) "" else "   (the perturbation)"),
         y = sprintf("log2FC  %s%s", resp_label,
                     if (is_control) "" else "   (the intervention)"))
  # The fitted slope is the statistic being reported, so draw it rather than
  # leaving the reader to infer it from the box.
  if (!is.null(st) && "tls_slope" %in% names(st) && is.finite(st$tls_slope[1])) {
    p <- p + geom_abline(slope = st$tls_slope[1], intercept = 0,
                         colour = "#E41A1C", linewidth = .8)
  }
  if (!is.null(box)) {
    p <- p + annotate("label", x = lim, y = lim, hjust = 1, vjust = 1, label = box,
                      size = 2.9, family = "mono", lineheight = 1.15,
                      fill = grDevices::adjustcolor("white", alpha.f = .88),
                      label.size = .25, colour = "grey15")
  }
  print(p)
}
```

> If a gene was pushed up by the knockout it should be pushed back down when reintroducing ATRX: points in the shaded
> quadrants, along the dashed line. Slope −1 is a complete rescue, 0 is no
> response. We see a slightly higher recovery in `ATRX_FL` compared to `ATRX_IFF`,
> however, both are very close to zero. Furthermore, using E6/EmptyVector vs FL/IFF as a control,
> we see that we obtain a steeper slope with a better $R^2$, while there should be no rescue there. 
> Therefore, we can conclude that there is no significant proof of rescue when introducing ATRX_FL.
> Any rescue effect could possibly be obfuscated by the large difference observed between the TP53 and E6 clones.

---

# The E6 lineage alone

With `TP53` excluded, every remaining group shares one clone and one
transduction step. Differences here are attributable to the construct — the two
effects we could not separate above are now held constant by design rather than
corrected for.

## PCA, E6-derived samples only

```{r e6-pca, fig.height = 5}
co <- grab("pca", "E6_derived", "no_vector_no_transgene_coords.csv")
ve <- grab("pca", "E6_derived", "no_vector_no_transgene_variance_explained.csv")
if (is.null(co)) {
  missing_note("pca", "E6_derived", "no_vector_no_transgene_coords.csv")
} else {
  pv <- if (!is.null(ve)) setNames(ve$percent_variance, ve$PC) else c(PC1=NA, PC2=NA)
  print(
    co %>% mutate(condition = cond_factor(condition)) %>%
      ggplot(aes(PC1, PC2, colour = condition, label = sample)) +
      geom_point(size = 3) +
      ggrepel::geom_text_repel(size = 2.6, show.legend = FALSE, max.overlaps = 20) +
      scale_colour_manual(values = PAL) +
      labs(title = "E6-derived samples, reporter and ATRX features removed",
           subtitle = "clone and transduction held constant; the construct is the only intended difference",
           x = sprintf("PC1: %.1f%% variance", pv[["PC1"]]),
           y = sprintf("PC2: %.1f%% variance", pv[["PC2"]]),
           colour = "Condition")
  )
}
```

## Sample–sample correlation, E6-derived only

Pairwise correlations are unchanged by dropping TP53 — but the colour scale
rescales to the range that remains, so differences compressed by the clonal
outlier become visible.

```{r e6-cor, fig.height = 5.5}
if (is.null(cm)) missing_note("design_qc", "sample_correlation.csv") else {
  m <- as.matrix(cm[, -1]); rownames(m) <- cm$sample
  keep <- meta$sample[meta$condition != "TP53" & meta$sample %in% rownames(m)]
  if (length(keep) < 3) {
    cat("\n> Fewer than 3 non-TP53 samples — nothing to plot.\n\n")
  } else {
    m <- m[keep, keep]
    print(
      as.data.frame(as.table(m)) %>%
        setNames(c("x", "y", "rho")) %>%
        mutate(x = factor(x, keep), y = factor(y, rev(keep))) %>%
        ggplot(aes(x, y, fill = rho)) +
        geom_tile(colour = "white", linewidth = 0.3) +
        geom_text(aes(label = sprintf("%.3f", rho)), size = 2.1, colour = "grey20") +
        scale_fill_gradient(low = "white", high = "#2166AC") +
        coord_fixed() +
        labs(title = "Spearman correlation (VST), TP53 excluded",
             subtitle = "scale rescaled to the E6 lineage",
             x = NULL, y = NULL) +
        theme(axis.text.x = element_text(angle = 45, hjust = 1),
              panel.grid = element_blank())
    )
  }
}
```

## The three construct comparisons

```{r e6-three, fig.height = 4.5, results = "asis"}
trio <- list(
  c("ATRX_FL_vs_EmptyVector",  "ATRX_FL",  "EmptyVector",
    "Full-length rescue against its own transduced control"),
  c("ATRX_IFF_vs_EmptyVector", "ATRX_IFF", "EmptyVector",
    "In-frame fusion against the same control")
)
trio_tally <- list()
for (t4 in trio) {
  r <- de_results("deseq2", t4[1], "results.csv")
  cat("\n\n###", t4[4], "\n\n")
  if (is.null(r)) { missing_note("deseq2", t4[1], "results.csv"); next }
  trio_tally[[t4[1]]] <- de_counts(r)
  print(volcano(r, t4[1], num = t4[2], den = t4[3]))
  cat("\n\n")
  print(kable(top_genes(r, n = 8)))
  cat("\n\n")
}
```

### The two constructs against each other

```{r e6-three-fl-iff-rna, fig.height = 4.5, results = "asis"}
r_fl_iff <- de_results("deseq2", "ATRX_IFF_vs_ATRX_FL", "results.csv")
if (is.null(r_fl_iff)) {
  missing_note("deseq2", "ATRX_IFF_vs_ATRX_FL", "results.csv")
} else {
  trio_tally[["ATRX_IFF_vs_ATRX_FL"]] <- de_counts(r_fl_iff)
  print(volcano(r_fl_iff, "ATRX_IFF_vs_ATRX_FL", num = "ATRX_IFF", den = "ATRX_FL"))
  cat("\n\n")
  print(kable(top_genes(r_fl_iff, n = 8)))
}
```

> _Add interpretation here._

```{r e6-three-fl-iff-prot, fig.height = 4.5, results = "asis"}
if (!is.null(r_fl_iff)) {
  prot_sig <- load_proteomics_sig()
  if (is.null(prot_sig)) {
    cat("\n> **Not available.** Proteomics overlay skipped: `params$proteomics_xlsx`",
        "was not found, or is missing the expected sheet/columns for comparison",
        paste0("`", params$proteomics_comparison, "`.\n\n"))
  } else {
    cat("\n\n**Proteomics concordance (RNA vs LC-MS/MS protein), same contrast:**\n\n")
    print(proteomic_volcano(r_fl_iff, "Proteomics-filtered: ATRX_IFF_vs_ATRX_FL",
                            num = "ATRX_IFF", den = "ATRX_FL", prot_sig))
  }
}
```

> _Add interpretation here._

```{r e6-trio-tally}
if (length(trio_tally)) {
  do.call(rbind, trio_tally) %>% as.data.frame() %>%
    tibble::rownames_to_column("contrast") %>% mutate(total = up + down) %>%
    kable(caption = "Differentially expressed genes within the E6 lineage")
}
```

## Genes of interest, E6 lineage only

```{r e6-goi, fig.height = 4}
# Size factors (and so normalised counts) are fit from all samples for every
# contrast, so any contrast's normalized_counts.csv carries the same values —
# reuse E6_vs_TP53, already grabbed above.
goi <- gene_long(grab("deseq2", "E6_vs_TP53", "normalized_counts.csv"), meta)
if (!is.null(goi)) {
  goi <- goi %>% filter(tolower(gene_id) %in% tolower(params$highlight_genes))
}
if (is.null(goi) || nrow(goi) == 0) {
  missing_note("deseq2", "E6_vs_TP53", "normalized_counts.csv")
} else {
  print(
    goi %>% filter(condition != "TP53") %>%
      group_by(gene_id, condition) %>%
      summarise(mean = mean(value), sd = sd(value), .groups = "drop") %>%
      ggplot(aes(condition, mean, fill = condition)) +
      geom_col() +
      geom_errorbar(aes(ymin = pmax(mean - sd, 0), ymax = mean + sd),
                    width = .25, linewidth = .3) +
      facet_wrap(~ gene_id, scales = "free_y", nrow = 1) +
      scale_fill_manual(values = PAL, guide = "none") +
      labs(title = "Genes of interest, TP53 excluded", x = NULL,
           y = "normalised count") +
      theme(axis.text.x = element_text(angle = 45, hjust = 1))
  )
}
```

> Here, we got more coherent results in both HOXB genes:
> taking E6 as baseline, we can see that HOXB expression is always lower in `ATRX_FL` 
> than `E6`, while it is much higher in `ATRX_IFF`. Interestingly, this could mean that `ATRX_IFF` has an acitve role in disturbing PRC2
> 

---

# Pathway enrichment (GSEA)

```{r gsea, fig.height = 6, results = "asis"}
gsea_dot <- function(df, title, top = 12) {
  df <- df %>% filter(!is.na(padj)) %>% arrange(desc(abs(NES)))
  df <- bind_rows(head(filter(df, NES > 0), top), head(filter(df, NES < 0), top))
  if (!nrow(df)) return(invisible(NULL))
  num <- df$numerator[1]; den <- df$denominator[1]
  up   <- sprintf("NES > 0: pathway genes HIGHER in %s", num)
  down <- sprintf("NES < 0: pathway genes HIGHER in %s", den)
  df <- df %>% mutate(
    pathway = gsub("^HALLMARK_", "", pathway),
    pathway = factor(pathway, levels = pathway[order(df$NES)]),
    side = factor(ifelse(NES > 0, up, down), levels = c(up, down)))
  print(
    ggplot(df, aes(NES, pathway, colour = padj, size = size)) +
      geom_point() +
      geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3) +
      facet_grid(side ~ ., scales = "free_y", space = "free_y") +
      scale_colour_gradient(low = "#B2182B", high = "grey75", limits = c(0, 0.25)) +
      labs(title = title,
           subtitle = sprintf("enrichment of the %s / %s ranking", num, den),
           y = NULL, colour = "padj", size = "Set size") +
      theme(strip.text.y = element_text(angle = 0, size = 8))
  )
}
# EDIT: which contrasts to show enrichment for.
for (cn in c("ATRX_IFF_vs_ATRX_FL", "ATRX_FL_vs_EmptyVector",
             "ATRX_IFF_vs_EmptyVector")) {
  g <- grab("gsea", cn, "H_results.csv")
  cat("\n\n##", cn, "\n\n")
  if (is.null(g)) { missing_note("gsea", cn, "H_results.csv"); next }
  gsea_dot(g, paste("Hallmark:", cn))
}
```

> Hallmark proliferation sets (E2F, G2M, MYC) are enriched in `ATRX_IFF` compared to `ATRX_FL`, and
> the inflammatory/interferon sets are enriched in `ATRX_FL`.

## KEGG pathway categories

The five KEGG BRITE categories, drawn from the live KEGG definitions rather than
MSigDB's frozen snapshot. Unlike the collection dotplot above, **every** pathway
in a category is shown: a category is a short fixed list, so the pathways that
did not move are part of the answer. Filled points are `padj < 0.05` and coloured
across that range; hollow grey points did not reach it. Restricted to the
headline `ATRX_IFF_vs_ATRX_FL` contrast — the cleanest comparison in the design.

```{r kegg-categories, fig.height = 4.5, results = "asis"}
KEGG_CATS <- c(replication_and_repair       = "Replication and repair",
               chromosome                   = "Chromosome",
               cell_growth_and_death        = "Cell growth and death",
               nervous_system               = "Nervous system",
               development_and_regeneration = "Development and regeneration")

kegg_cat_dot <- function(df, title) {
  if (is.null(df) || !nrow(df)) return(invisible(NULL))
  cut <- params$padj
  df$significant <- !is.na(df$padj) & df$padj < cut
  df$pathway <- factor(df$pathway, levels = df$pathway[order(df$NES)])
  sz <- range(df$size, na.rm = TRUE)
  brk <- if (sz[1] == sz[2]) sz[1] else unique(round(c(sz[1], mean(sz), sz[2])))
  # Spell the direction out the same way the Hallmark dotplot above does.
  num <- if ("numerator" %in% names(df)) df$numerator[1] else ""
  den <- if ("denominator" %in% names(df)) df$denominator[1] else ""
  sub <- if (nzchar(num) && nzchar(den))
    sprintf("NES > 0: pathway genes HIGHER in %s   |   NES < 0: pathway genes HIGHER in %s",
            num, den) else NULL
  print(
    ggplot(df, aes(NES, pathway)) +
      geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.3) +
      geom_point(data = df[!df$significant, ], aes(size = size),
                 shape = 21, colour = "grey65", fill = NA, stroke = 0.7) +
      geom_point(data = df[df$significant, ], aes(size = size, colour = padj),
                 shape = 19) +
      scale_colour_gradient(low = "#A50F15", high = "#FC9272",
                            limits = c(0, cut), oob = scales::squish) +
      scale_size_continuous(limits = sz, breaks = brk, range = c(2, 8)) +
      labs(title = title, subtitle = sub, y = NULL, colour = "padj", size = "Set size") +
      theme(plot.subtitle = element_text(size = 8, colour = "grey30"))
  )
}

# EDIT: which contrast to show category panels for. Headline contrast only.
cn <- "ATRX_IFF_vs_ATRX_FL"
for (slug in names(KEGG_CATS)) {
  g <- grab("gsea", cn, "kegg", paste0(slug, "_results.csv"))
  cat("\n\n###", KEGG_CATS[[slug]], "\n\n")
  if (is.null(g)) { missing_note("gsea", cn, "kegg", paste0(slug, "_results.csv")); next }
  kegg_cat_dot(g, paste(KEGG_CATS[[slug]], "-", cn))
}
```

---

# Summary

TBD

```{r session, results = "asis"}
cat("\n<details><summary>Session info</summary>\n\n```\n")
cat(capture.output(sessionInfo()), sep = "\n")
cat("\n```\n</details>\n")
```
